---
title: "Final Project for STAR Project"
author: "Ruitian Liu, Zhiye Jiang, Tommy Yu, Yuyu Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
       collapsed: true
       smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    fig_width: 8
    fig_height: 5
    fig_caption: true
fontsize: 11pt
bibliography: references.bib
geometry: margin=1in
---


```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Lora:ital,wght@0,400;0,600;1,400&family=Source+Code+Pro:wght@400;600&family=Inter:wght@300;400;500;600&display=swap');

body {
  font-family: 'Inter', sans-serif;
  font-size: 15px;
  line-height: 1.85;
  color: #2d2d2d;
  background-color: #fafaf8;
}


h1, h2, h3, h4 {
  font-family: 'Lora', Georgia, serif;
  font-weight: 600;
  letter-spacing: -0.02em;
  color: #1a1a1a;
}

h1.title {
  font-size: 2.4em;
  border-bottom: 3px solid #2563eb;
  padding-bottom: 0.3em;
  margin-bottom: 0.2em;
}

h1 { font-size: 1.9em; border-left: 4px solid #2563eb; padding-left: 0.6em; }
h2 { font-size: 1.45em; color: #374151; border-bottom: 1px solid #e5e7eb; padding-bottom: 0.2em; }
h3 { font-size: 1.15em; color: #4b5563; }

.author, .date {
  font-family: 'Inter', sans-serif;
  font-size: 0.95em;
  color: #6b7280;
  font-weight: 400;
}


p { margin-bottom: 1.1em; max-width: 760px; text-align: justify;}


pre, code {
  font-family: 'Source Code Pro', 'Courier New', monospace;
  font-size: 0.88em;
}

pre {
  background-color: #f1f5f9;
  border: 1px solid #e2e8f0;
  border-left: 4px solid #2563eb;
  border-radius: 6px;
  padding: 1em 1.2em;
}


code:not(pre code) {
  background: #eff6ff;
  color: #1d4ed8;
  padding: 0.15em 0.4em;
  border-radius: 3px;
  font-size: 0.85em;
}


table {
  border-collapse: collapse;
  width: 100%;
  margin: 1.5em 0;
  font-size: 0.92em;
}

thead tr { background-color: #1e3a5f; color: #ffffff; }
th { padding: 10px 14px; text-align: left; font-weight: 600; }
td { padding: 9px 14px; border-bottom: 1px solid #e5e7eb; }
tbody tr:nth-child(even) { background-color: #f8fafc; }
tbody tr:hover { background-color: #eff6ff; transition: background 0.15s; }


blockquote {
  border-left: 4px solid #93c5fd;
  background: #eff6ff;
  margin: 1.5em 0;
  padding: 0.8em 1.2em;
  border-radius: 0 6px 6px 0;
  color: #1e40af;
  font-style: italic;
}

#TOC {
  font-size: 0.88em;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  padding: 1em;
  background: #ffffff;
  box-shadow: 0 1px 4px rgba(0,0,0,0.06);

  /* 限制高度，超出后滚动 */
  max-height: calc(100vh - 100px);
  overflow-y: auto;

  /* 自定义滚动条（Webkit/Chrome/Edge） */
  scrollbar-width: thin;                        /* Firefox */
  scrollbar-color: #cbd5e1 transparent;         /* Firefox: thumb / track */
}

/* Chrome / Safari / Edge */
#TOC::-webkit-scrollbar {
  width: 5px;
}
#TOC::-webkit-scrollbar-track {
  background: transparent;
}
#TOC::-webkit-scrollbar-thumb {
  background-color: #cbd5e1;
  border-radius: 99px;
}
#TOC::-webkit-scrollbar-thumb:hover {
  background-color: #2563eb;
}

img { border-radius: 6px; box-shadow: 0 2px 12px rgba(0,0,0,0.08); }

hr { border: none; border-top: 2px solid #e5e7eb; margin: 2em 0; }

.main-container { max-width: 960px; padding-left: 60px; padding-right: 60px;}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
pretty_label <- function(x) {
  x <- as.character(x)
  dict <- c(
    area_type = "Area Type",
    class_type = "Class Type",
    teacher_degree = "Teacher's Degree",
    teacher_career = "Teacher's Career Stage",
    student_gender = "Student's Gender",
    teacher_gender = "Teacher's Gender",
    free_lunch = "Free Lunch Status",
    teacher_race = "Teacher's Race",
    student_race = "Student's Race",
    math_score = "Math Score",
    "class_type * grade" = "Class × Grade"
  )
  out <- ifelse(x %in% names(dict), dict[x], x)
  out <- gsub("_", " ", out)
  out <- trimws(out)
  out <- sapply(strsplit(out, " +"), function(w) {
    paste0(toupper(substring(w, 1, 1)), substring(w, 2), collapse = " ")
  })
  unname(out)
}

fmt_p <- function(p, digits = 3, threshold = 1e-3) {
  if (is.na(p)) return(NA_character_)
  if (p < threshold) return(paste0("< ", formatC(threshold, format = "f", digits = digits)))
  formatC(p, format = "f", digits = digits)
}

options(warn = -1)
library(tidyverse)
library(tableone)
library(mice)
library(lme4)
library(gt)
library(patchwork)
library(stringr)
library(dplyr)
library(tableone)
library(tidyr)
library(patchwork)
library(purrr)
library(rcompanion)
library(psych)
library(ggplot2)
library(dlookr)
library(modelsummary)
library(broom)
library(forcats)
library(ggplot2)
```

# Abstract

# Background and study design

Project STAR (Student/Teacher Achievement Ratio) was one of the most important large-scale randomized education experiments in the United States. It was launched by the Tennessee State Department of Education in the 1985–86 school year to study whether smaller classes in the early elementary grades improve student outcomes. The project began in 79 schools across 42 districts in Tennessee and eventually followed students over many years. The large sample and long follow-up period had made STAR an important study in education fields.

In the STAR experiment, students entering kindergarten were randomly assigned within schools to one of three classroom types: small classes (about 13–17 students), regular classes (about 22–25 students), and regular classes with a full-time teacher’s aide. Teachers were also randomly assigned to these class types. The original design intended for students to stay in their assigned treatment group after entering the study. 

Class size has long been a question for educators. **Smaller classes** may help young children by giving them more individual attention and improving classroom participation. The potential benefits of class size may be especially important in the **early grades**. However, reducing class size is expensive, since it requires more teachers, more classrooms and funding. This is a big challenge for underserved schools or districts, and had become an important equity concern in education. For this reason, a question arises: **whether the gains of small classes are worthy to justify the substantial costs of implementation**. [@schanzenbach2007star]

Students were intended to stay in the same class-type assignment throughout the whole process. So the study was began with randomization, and all assignments was conducted within schools, the design helped balance school-level differences across treatment groups. This can be a good background for interpreting casual effects.

Although STAR began with randomization, later studies have claimed that it is **not a completely randomized experiments**. Because the experiments are affected by factors such as student transfers, classroom reassignment, imperfect compliance, teacher changes, and missing follow-up data. When missing of the data is not at random, it can generate selection bias. Therefore, studies must address these post-randomization outcomes carefully.


## Some empirical complications violating randomization

- **Student mobility / transfers.**

@chetty2011classroom claimed that some students did not remain in their originally assigned classrooms or schools over time, and new entrants were also added in later grades. As a result, treatment exposure in later grades may differ from the initial assignment.

- **Noncompliance with original assignment.**

@krueger1999star shown that some students moved between class types after the initial random assignment. Therefore, observed class type in later grades would not always be consistent with original treatment assignment.

- **Teacher changes and classroom reshuffling.**

@chetty2011classroom shown that if students were exposed to different classroom environments over time, including new teachers in subsequent grades and changes in class composition. These post-randomization changes may affect achievement independently of the original class-size assignment.

- **Peer Effect.**

@cascio2007first argued that scores may depend also on the characteristics of classmates. Since classroom composition can affect learning outcomes, peer effects may operate alongside the direct effect of small classes and make interpretation of the treatment difficult.

## Important Violations of Randomization - Selection Bias

One persistent concern of STAR Project is that there was substantial sample loss in the later years of the study, and this attrition may not have been entirely random. More broadly, @hanushek1999star identified attrition, crossovers, and selective test taking as threats to the integrity of STAR’s original randomization, and @filges2018classsize note that these remain central concerns in later assessments of the experiment.

- **Attrition is related to academic performance or student disadvantage.**

@ding2006dynamic shown that attrition in Project STAR was not random with respect to student performance. **Students who later attrited had significantly lower kindergarten achievement**, and that among students on free lunch, those who left the sample scored significantly **lower in mathematics** than comparable students who remained. In terms of this, if lower-performing students are more likely to leave the sample, the observed students may no longer represent randomization.

- **Attrition is also related to treatment status.**

@krueger1999star shown that attrition from STAR was substantial. In a subsample with initial enrollment logs, 10.4% of students assigned to small classes were missing by the start of kindergarten, compared with 14.3% of those assigned to regular classes, suggesting that some students may have been withdrawn because they were not assigned to a small class. In addition, @ding2006dynamic show that the pattern of **attrition differed between class types within and across schools**. These findings supported that the observed small-class and regular-class groups in later grades are no longer simply the originally randomized groups, but the outcome of initial random assignment followed by differential attrition and movement across class types.

# Research Questions

## Assumptions

- **Stable Unit Treatment Value Assumption (SUTVA):**

A first assumption is no interference. Each classroom’s potential outcome is assumed to be independent of the treatment assigned to other classrooms. The second assumption is no hidden versions of treatment. Small class should not contain multiple hidden versions of treatment.

- **Positivity:**

Within the relevant strata of background characteristics such as grade, race, free lunch status, and school context, there should be sufficient overlap so that no subgroup is deterministically assigned to only one class type.

- **Conditional ignorability of attrition:**

Conditional on the observed baseline covariates, treatment status, prior achievement, and other auxiliary variables related to student performance and participation, whether a student remains observed is assumed to be independent of the missing potential math outcomes.

- **Random assignment at baseline:**

At study entry, students were randomly assigned within schools to class types, so baseline treatment assignment can be treated as independent of students’ pre-treatment characteristics and potential outcomes.

- **Exclusive Restriction:**

Conditional on the class type actually received, the original random assignment affects math achievement only through treatment receipt, not through any direct channel such as parental responses, administrative reassignment, compliance behavior, etc.

## Questions of Interests

- How robust are the estimated treatment effects under multiple missing-data imputation methods?

- Whether missingness in math outcomes can be explained by observed background and prior-achievement variables, making a MAR-type assumption more plausible.

- Whether students assigned to smaller classes achieve higher mathematics scores than those in regular classes and whether this relationship differs across grades K--3 after resolving selection bias.

# Exploratory Data Analysis

```{r load-data}
# Please ensure the file exists at this path
df <- read.table("STAR.tab", header = TRUE, sep = "\t")
```

  
```{r phase-1}
## Phase 1: Column Name Filtering and Cleanin
# Truncate column names to the first 8 characters; STAR dataset column names usually contain grade prefixes (e.g., gk, g1, etc.)
names(df) <- substr(names(df), 1, 8)
cols <- names(df)

# Use regular expressions to extract column names starting with gk, g1, g2, g3
m <- regexpr("^(gk|g1|g2|g3)(.*)$", cols, perl = TRUE)
prefix <- regmatches(cols, m)
suffix <- sub("^(gk|g1|g2|g3)", "", prefix)

# Count the frequency of each variable and filter variables that exist across all grades (appearing at least 4 times)
tab <- table(suffix)
valid_suffix <- names(tab[tab >= 4])

# Construct the list of column names to keep
keep_cols <- unlist(lapply(valid_suffix, function(s) paste0(c("gk","g1","g2","g3"), s)))

# Separate basic student information (assuming the first 6 columns are basic info) and teaching data corresponding to grades
student_info <- df[, 1:6, drop = FALSE]
df_keep <- df[, intersect(names(df), keep_cols), drop = FALSE]
star <- cbind(student_info, df_keep)
```

  
```{r phase-2}
## Phase 2: Data Reshaping (Wide to Long -> Long to Wide)
# Convert the dataset from wide format to long format, then back to wide format, to standardize the grade structure
stars <- star %>%
  pivot_longer(cols = matches("^(gk|g1|g2|g3)"),
               names_to = c("grade", "variable"),
               names_pattern = "^g([k123])(.*)$", # Capture grade (k, 1, 2, 3) and variable names
               values_to = "value") %>%
  pivot_wider(names_from = variable,
              values_from = value) %>%
  # Convert the grade column into an ordered factor
  mutate(grade = dplyr::recode(grade, k = "K"),
         grade = factor(grade, levels = c("K", "1", "2", "3"), ordered = FALSE))

```


## Data Cleaning and Preprocessing

The dataset downloaded from Harvard Dataverse was first examined. It contains 11,601 observations and 379 variables, with a total of 2,477,601 missing entries. Given its large size and substantial amount of missing data, it is necessary to retain only variables relevant to the analysis and to address missingness carefully.

```{r phase-3}
## Phase 3: Renaming and Variable Type Conversion
# Standardize variable names to make them more readable
stars <- stars %>% rename(
  student_id = stdntid, birth_month = birthmon, birth_day = birthday, 
  birth_year = birthyea, teaching_years = tyears, class_size = classs, 
  math_score = tmaths, school_id = schid, teacher_id = tchid)
stars_diag <- stars %>% rename(area_type = surban, class_type = classt, teacher_degree = thighd,
    teacher_career = tcaree, student_gender = gender, teacher_gender = tgen,
    free_lunch = freelu, teacher_race = trace, student_race = race)

stars_diag[c("area_type","class_type","teacher_degree","teacher_career",
             "student_gender","teacher_gender","free_lunch",
             "teacher_race","student_race")] <- 
  lapply(stars_diag[c("area_type","class_type","teacher_degree","teacher_career",
                      "student_gender","teacher_gender","free_lunch",
                      "teacher_race","student_race")], factor)

stars_diag %>% dplyr::filter(if_all(where(is.factor), ~ !is.na(.))) %>% diagnose_category()
```

  
Following the Project STAR codebook, we recoded several categorical variables and collapsed categories when the original coding was too sparse for stable analysis. In particular, student race and teacher race were simplified to Black vs. Non-Black, since the STAR documentation shows that racial categories other than White and Black were extremely rare in the data. Teacher degree was reduced to Below Master versus Master or above, because specialist and doctoral categories were infrequent relative to the dominant bachelor’s/master’s groups. Teacher career stage was also collapsed into Lower level and Upper level, since the original multi-level ladder structure was highly imbalanced in practice.


```{r phase-4}
## Phase 4: Feature Engineering and Variable Cleaning
# Use case_when to create more descriptive categorical variables and delete unnecessary raw variables
stars1 <- stars %>% mutate(
  # Area type encoding conversion
  area_type = factor(case_when(
    surban == 1 ~ "Inner City",
    surban == 2 ~ "Suburban",
    surban == 3 ~ "Rural",
    surban == 4 ~ "Urban",
    TRUE ~ NA_character_)),
  # Class type encoding conversion
  class_type = factor(classt, levels = c(1, 2, 3),
                      labels = c("Small class", "Regular class", "Regular + aide")),
  # Teacher degree level binning
  teacher_degree = factor(case_when(
    thighd %in% c(1, 2) ~ "Below Master",
    thighd %in% c(3, 4, 5, 6) ~ "Master or above",
    TRUE ~ NA_character_),
    levels = c("Below Master", "Master or above"), ordered = TRUE),
  # Teacher career stage binning
  teacher_career = factor(case_when(
    tcaree %in% c(1, 2, 3) ~ "Lower level",
    tcaree %in% c(4, 5, 6, 7) ~ "Upper level",
    TRUE ~ NA_character_),
    levels = c("Lower level", "Upper level"), ordered = TRUE),
  # Gender and free lunch encoding conversion
  student_gender = factor(gender, levels = c(1, 2),
                          labels = c("Male", "Female")),
  teacher_gender = factor(tgen, levels = c(1, 2),
                          labels = c("Male", "Female")),
  free_lunch = factor(freelu, levels = c(1, 2),
                      labels = c("Free lunch", "Non-free lunch")),
  # Racial binarization
  teacher_race = factor(case_when(
    trace == 2 ~ "Black",
    trace %in% c(1, 3, 4, 5, 6) ~ "Non-Black",
    TRUE ~ NA_character_),
    levels = c("Black", "Non-Black")),
  student_race = factor(case_when(
    race == 2 ~ "Black",
    race %in% c(1, 3, 4, 5, 6) ~ "Non-Black",
    TRUE ~ NA_character_),
    levels = c("Black", "Non-Black"))
) %>%
  # Delete the converted raw variables and analysis indicators that are no longer needed; here, we only keep the math scores for performance variables
  select(-surban, -classt, -thighd, -gender, -race, -tcaree, -tgen, -freelu, -trace, -starts_with("gk"))
```


## Identification of potential confounder

Let the variables be defined as follows:

- $Y$: Outcome variable — students’ **math scores**.
- $X$: Treatment variables — **class type, grade level, and their interaction (class type × grade)**.
- $Z$: Potential confounders — **school area type, free lunch status, student race, teacher race, and teacher degree, etc**.
- $A$: Ancillary variables — **treads (reading), tlists (listening), wordsk (word skills), motivr (motivation), selfco (self-concept), and lagged math scores from previous grades**.

Our objective is to estimate the effect of class size on students’ mathematics achievement in the STAR experiment. Although the initial assignment was randomized, factors such as classroom reassignment and student mobility may weaken the ideal experimental conditions. As a result, background characteristics may influence both classroom assignment and student outcomes.

A variable $Z$ acts as a confounder if it is associated with both the treatment $X$ (class type) and the outcome $Y$ (math score). Thus we examine three relationships:

- **$X \rightarrow Y$:** whether class type is associated with math achievement.  
- **$Z \rightarrow Y$:** whether background characteristics are associated with math achievement.
- **$Z \rightarrow X$ or $X \leftarrow Z \rightarrow Y$:** whether background characteristics are related to classroom assignment. 

If these relationships hold, failing to control for $Z$ may bias estimates of the effect of class size on student outcomes. Therefore, these variables are included as controls in the subsequent analysis.


### Verify **$X \rightarrow Y$:**

```{r, fig.width=6, fig.height=4, out.width = "100%" }
# Plots and Graphs
library(ggplot2)
stars1 <- stars1 %>%
  mutate(class_type = case_when(
    class_type %in% c("Regular class", "Regular + aide") ~ "Regular",
    TRUE ~ class_type
  )) %>%
  mutate(class_type = as.factor(class_type))
ggplot(data=stars1, mapping=aes(y=math_score,x=grade,fill=class_type)) +
   geom_boxplot(na.rm=TRUE) + 
   xlab('Grade Level')+ ylab('Math Scores') + 
   ggtitle('Math Scores by Level and Class Type') +
   theme_bw(base_size = 16) +
   theme(plot.title = element_text(size = 18, hjust = 0.5),
         axis.title = element_text(size = 16),
         axis.text = element_text(size = 14),
         legend.text = element_text(size = 14),
         legend.title = element_text(size = 15))
```

The boxplots show the distribution of students’ mathematics scores by grade level (K–3) and class type (small class/regular class). We can see that math scores increase steadily across grades. Also, students in small classes consistently exhibit slightly higher math scores than those in regular classes within the same grade. Thirdly, the difference between small classes and regular classes appears to persist across all grade levels, suggesting a potential interaction between grade and class type. In particular, the gap between the two class types appears to be slightly larger in lower grades. Thus, we can say that $X \rightarrow Y$ exists.


### Examine potential variables that **$Z \rightarrow Y$:**

```{r, fig.width=6, fig.height=4, out.width = "100%" }
library(patchwork)
options(warn = -1)
cat_vars <- c("area_type", "class_type", "teacher_degree", "teacher_career",
              "student_gender", "teacher_gender", "free_lunch", "teacher_race", "student_race")
library(ggdist)
make_all_plots_violin <- function(df, cont, cats, ncol = 3) {
  cont_lab <- pretty_label(cont)
  one_cat <- function(cat_var) {
    d <- dplyr::filter(df, !is.na(.data[[cat_var]]), !is.na(.data[[cont]]))
    if (nrow(d) == 0) return(NULL)
    p_txt <- fmt_p(kruskal.test(as.formula(paste(cont, "~", cat_var)), data = d)$p.value)
    y <- quantile(d[[cont]], c(.01, .99), na.rm = TRUE)
    ggplot2::ggplot(d, ggplot2::aes(.data[[cat_var]], .data[[cont]], fill = .data[[cat_var]])) +
      ggplot2::geom_violin(trim = TRUE, alpha = 0.7, color = NA) +
      ggplot2::geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", color = "grey30", alpha = 0.8) +
      ggplot2::stat_summary(fun = mean, geom = "point", shape = 18, size = 2.5, color = "red") +
      ggplot2::coord_cartesian(ylim = y) +
      ggplot2::labs(title = paste0(pretty_label(cat_var), " (p = ", p_txt, ")"), x = NULL, y = cont_lab) +
      ggplot2::theme_bw(base_size = 13) +
      ggplot2::theme(legend.position = "none",
        plot.title = ggplot2::element_text(size = 13),
        axis.title = ggplot2::element_text(size = 12),
        axis.text = ggplot2::element_text(size = 11))
  }
  plots <- Filter(Negate(is.null), lapply(cats, one_cat))
  patchwork::wrap_plots(plots, ncol = ncol) +
    patchwork::plot_annotation(
      title = paste0("Distributions of ", cont_lab, " under different settings"),
      subtitle = "Kruskal-Wallis p-values in panel titles | Red diamond = mean",
      theme = ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
        plot.subtitle = ggplot2::element_text(hjust = 0.5)))
}

make_all_plots_violin(stars1, "math_score", cat_vars)
```

The figure compares math score distributions across several student, teacher, and school characteristics. As we can see, some of the distribution of data has a strong violations of normality. So we use Kruskal--Wallis test to evaluate whether math scores differ across groups. Unlike the ANOVA, Kruskal--Wallis it is a nonparametric method based on ranks, which means that there are no normality assumption of data.

The null hypothesis of Kruskal--Wallis is that the distributions of math scores are the same across all groups; equivalently, under a common-shape assumption, the group medians are equal. The alternative hypothesis is that at least one group differs from the others.

Because multiple tests are conducted across several covariates, we adopt a significance threshold of $\alpha =0.001$ to reduce the risk of false positives arising from multiple comparisons and the large sample size of the STAR dataset. The figure compares math score distributions across several student, teacher, and school characteristics. 

In this case, Kruksal--Wallis p--value of **area type, free lunch status, student race, teacher race, teacher gender, and teacher degree** are less or equal than 0.001. Thus, we reject the null hypothesis, showing that they are significant factors that influence students’ mathematics performance, indicating their relationship of $Z \rightarrow Y$. To determine whether they act as confounders, we further examine whether these variables are also associated with class type (or to say $Z \rightarrow X$ exists).


### Examine confounders that satistfies **$Z \leftarrow X \rightarrow Y$:**

```{r, fig.width=6, fig.height=4, out.width = "100%" }

all_vars <- c("math_score", "class_type", "grade", "free_lunch", 
              "student_race", "area_type", "teacher_degree", "teacher_gender")

stars_clean <- stars1[complete.cases(stars1[, all_vars]), ]

model_1 <- lm(math_score ~ class_type * grade, data = stars_clean)
model_2 <- lm(math_score ~ class_type * grade + free_lunch, data = stars_clean)
model_3 <- lm(math_score ~ class_type * grade + free_lunch + area_type, data = stars_clean)
model_4 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type, data = stars_clean)
model_5 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type + teacher_degree, data = stars_clean)
model_6 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type + teacher_degree + teacher_gender, data = stars_clean)

models <- list("M1" = model_1, "M2" = model_2, "M3" = model_3, "M4" = model_4, "M5" = model_5, "M6" = model_6)

model_note <- c("M1: Baseline", "M2: + Free lunch", "M3: + Student race",
  "M4: + Area type", "M5: + Teacher degree", "M6: + Teacher gender")

keep_terms <- c("class_typeSmall class", "class_typeSmall class:grade1",
  "class_typeSmall class:grade2", "class_typeSmall class:grade3",
  "grade1", "grade2", "grade3")

term_labels <- c(
  "class_typeSmall class" = "Small class",
  "class_typeSmall class:grade1" = "Grade 1 × Small class",
  "class_typeSmall class:grade2" = "Grade 2 × Small class",
  "class_typeSmall class:grade3" = "Grade 3 × Small class",
  "grade1" = "Grade 1",
  "grade2" = "Grade 2",
  "grade3" = "Grade 3"
)

coef_df <- imap_dfr(models, ~{
  tidy(.x, conf.int = TRUE) %>%
    mutate(model = .y)
}) %>%
  filter(term %in% keep_terms) %>%
  mutate(term_label = recode(term, !!!term_labels),
    term_group = case_when(term == "class_typeSmall class" ~ "Treatment main effect",
      str_detect(term, "class_typeSmall class:") ~ "Treatment × grade",
      str_detect(term, "^grade") ~ "Grade main effect"),
    model = factor(model, levels = c("M1", "M2", "M3", "M4", "M5", "M6"))) %>%
  group_by(term) %>%
  arrange(model, .by_group = TRUE) %>%
  mutate(pct_change_prev = 100 * (estimate - lag(estimate)) / abs(lag(estimate)),
    label_change = case_when(
      row_number() == 1 ~ "Base",
      is.infinite(pct_change_prev) | is.nan(pct_change_prev) ~ "",
      TRUE ~ sprintf("%+.1f%%", pct_change_prev))) %>%
  ungroup() %>%
  mutate(model = fct_rev(model))


coef_df <- coef_df %>%
  group_by(term_label) %>%
  mutate(xmin = min(conf.low), xmax = max(conf.high), xrange = xmax - xmin, label_x = estimate + 0.05 * xrange) %>% ungroup()


p <- ggplot(coef_df, aes(x = estimate, y = model, color = term_group)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 2.4) +
  geom_text(aes(x = label_x, label = label_change), size = 4, hjust = 0, color = "black") +
  facet_wrap(~ term_label, scales = "free_x", ncol = 3) +
   labs(title = "Coefficient stability across nested models",
    subtitle = paste("Baseline model (M1): Math Score ~ Class Type * Grade",
      "Labels show % change relative to the previous model", sep = "\n"),
    x = "Coefficient estimate (95% CI)", y = NULL,
    caption = paste("M1: Class Type * Grade", "M2: + Free Lunch Status", "M3: + Student's Race",
      "M4: + Area Type", "M5: + Teacher's Degree", "M6: + Teacher's Gender", sep = "   |   "))+
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 13),
    plot.caption = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 13)) +
  coord_cartesian(clip = "off")
p
```

Equivalent to the linear regression model

$$
Y_i=\alpha+X_i^\top\beta+Z_i^\top\gamma+\varepsilon_i,
$$

where $Y_i$ is math score, $X_i$ contains the treatment-related design terms for class type, grade, and their interaction terms, and $Z_i$ contains candidate confounders added sequentially across nested models. The figure tracks how the estimated coefficients $\beta$ change as additional covariates $Z_i$ are added step by step into the regression model. Each panel corresponds to one coefficient, each point is the estimate from a given model, and each horizontal line is its 95% confidence interval. The percentages show how much the coefficient changes relative to the previous model. 

The goal is to assess whether adding a covariate substantially changes the estimated effects of treatment variables. If a coefficient shifts noticeably after a variable is added, that variable may be acting as a confounder. As a heuristic, a covariate is flagged as a potential confounder if its inclusion changes a focal class-type coefficient by more than 10% in absolute value.

In this plot, the largest changes occur after adding free lunch status, area type and student’s race, while later adding teacher's degree and gender does not changes the coefficient of treatment effects. This suggests that these three variables account for the main confounding in the relationship of $Z \rightarrow X$ or adding $Z$ changes how $X \rightarrow Y$. Thus, in our final model we need to conditioning on the effects of free lunch status, area type and student’s race in order to accurately estimate the effect of treatment variable.


## Analysis of missing mechanism

- **MCAR (Missing Completely at Random)**: the probability of missingness is unrelated to both observed and unobserved data. In this case, missing values occur purely by chance, dropping incomplete cases does not introduce bias.

- **MAR (Missing at Random)**: the probability of missingness may depend on observed variables, but conditional on those observed variables, it does not depend on the missing value itself.

- **MNAR (Missing Not at Random)**: the probability of missingness still depends on the unobserved value itself, even after conditioning on observed data. This cannot be verified directly from the observed data alone, because the mechanism depends on values that are not observed.

In practice, these mechanisms cannot usually be distinguished with certainty from the observed data alone, especially in distinguishing MAR and MNAR. Therefore, rather than attempting to verify MAR directly, we assess its plausibility indirectly using the following diagnostic strategies:

- **Assess whether missingness is explainable by observed data**: model the missingness indicator using baseline covariates, lagged outcomes, and other observed variables. If missingness can be substantially explained by observed information, the MAR assumption becomes more plausible. [@rubin1976inference]

- **Include strong auxiliary variables**: add variables that are highly predictive of the missing outcome or of the missingness process into the imputation or likelihood-based model. This can reduce bias and strengthen the plausibility of MAR. [@collins2001comparison]; [@white2011multiple]

- **Check robustness under MAR-based methods**: compare results across approaches such as complete-case analysis, IPW, FIML, and MI under the same observed-data structure. Similar conclusions make MAR-based inference more credible. [@baraldi2010introduction]

- **Conduct MNAR sensitivity analysis**: evaluate how conclusions change under plausible departures from MAR using methods such as pattern-mixture models, selection models, or delta-adjusted imputation. Stable results suggest that the main findings are not overly sensitive to violations of MAR. [@daniels2000simple]


### Proving data is not MCAR

Each student’s outcome record is summarized by a four-digit indicator over grades K, 1, 2, 3, where 1 denotes an observed math score and 0 denotes a missing math score. For example, 1111 indicates complete observation across all four grades and 1001 indicates a non-monotone pattern with observation in K and Grade 3 but missingness in between.

We classified the K–3 observation patterns into five descriptive groups:

- Pattern 1111 was defined as **complete cases**. 
- Patterns 1110, 1100, 1000, 0100, and 0110 were grouped as **monotone attrition**, since once students entered a missing state, they did not subsequently return to full follow-up. 
- Patterns 0011 and 0001 were classified as **late entry**, as observations only began in later grades.
- Patterns 1010, 1001, 0101, 0010, and 1011 were classified as **intermittent**, reflecting non-monotone missingness with reappearance after one or more missing waves. 
- Pattern 0111 was separated as **no records in K**, because it there are a specific group of students who hasn't got kindergarten but observed continuously from Grade 1 onward.

```{r, fig.width=6, fig.height=4, out.width = "100%" }
wide <- stars1 %>% mutate(obs = !is.na(math_score)) %>%
  select(student_id, grade, obs) %>%
  pivot_wider(names_from = grade, values_from = obs)

wide <- wide %>% mutate(K = ifelse(is.na(K), FALSE, K),
    `1` = ifelse(is.na(`1`), FALSE, `1`),
    `2` = ifelse(is.na(`2`), FALSE, `2`),
    `3` = ifelse(is.na(`3`), FALSE, `3`))

wide <- wide %>% mutate(pattern = paste0(K*1, `1`*1, `2`*1, `3`*1))

wide <- wide %>% filter(pattern != "0000") %>%
  mutate(complete = ifelse(pattern == "1111", "Complete", "Incomplete"))

stars_plot <- stars1 %>% left_join(wide %>% select(student_id, pattern), by = "student_id") %>%
  filter(!is.na(pattern), pattern != "0000", !is.na(math_score))

pattern_order <- stars_plot %>% distinct(student_id, pattern) %>% count(pattern, sort = TRUE) %>% pull(pattern)
pattern_order <- c("1111", setdiff(pattern_order, "1111"))

stars_plot2 <- stars1 %>%
  left_join(wide %>% select(student_id, pattern), by = "student_id") %>%
  filter(!is.na(pattern), pattern != "0000", !is.na(math_score)) %>%
  mutate(pattern_class = case_when(pattern == "1111" ~ "Complete cases",
      pattern %in% c("1110", "1100", "1000", "0100", "0110") ~ "Monotone attrition",
      pattern %in% c("0011", "0001") ~ "Late entry",
      pattern %in% c("1010", "1001", "0101", "0010", "1011") ~ "Intermittent",
      pattern %in% "0111" ~ "No records in K",
      TRUE ~ NA_character_)) %>%
  filter(!is.na(pattern_class))

class_count <- stars_plot2 %>% distinct(student_id, pattern_class) %>% count(pattern_class)

class_labels <- class_count %>% mutate(label = paste0(pattern_class, "\n(n=", n, ")")) %>%
  {setNames(.$label, .$pattern_class)}

stars_plot2 <- stars_plot2 %>%
  mutate(pattern_class = factor(pattern_class, 
                                levels = c("Complete cases","Monotone attrition", 
                                           "Late entry", "Intermittent", "No records in K")))

p3 <- ggplot(stars_plot2, aes(x = pattern_class, y = math_score, fill = pattern_class)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_x_discrete(labels = class_labels) +
  labs(title = "Math score distribution by missingness type",
    subtitle = "Scores pooled across grades K–3",
    x = "Missingness type",
    y = "Math score") +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 13),
    axis.title = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "none")
p3 
``` 

The boxplot shows clear differences in pooled math-score distributions across missingness types from kindergarten to Grade 3. The pooled math-score distributions suggest that complete cases, intermittent cases, have broadly similar achievement levels. However, **the monotone attrition group has lower math score while the late-entry group shows significant higher math score among all categories**. The pattern suggests that missingness type is related to student performance, which is inconsistent with a purely MCAR mechanism. Thus, missingness type itself is informative and should be analyzed specifically.

### Examine whether missingness can be explained by observed variables

```{r}
baseline_vars <- c("math_score", "free_lunch", "student_race",
  "student_gender", "class_type", "area_type","class_size")

stu_base <- stars1 %>% arrange(student_id, grade) %>%
  group_by(student_id) %>% slice(1) %>%
  ungroup() %>%
  select(student_id, all_of(baseline_vars)) %>%
  left_join(wide %>% select(student_id, complete), by = "student_id")

tab <- CreateTableOne(vars = baseline_vars, strata = "complete",
  data = stu_base, test = TRUE)
tab_mat <- print(tab, smd = TRUE, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
tab_df  <- as.data.frame(tab_mat)
tab_df$Variable <- c("n", "Math score, mean (SD)", "Free lunch: Non-free lunch, n (%)",
  "Student race: Non-Black, n (%)", "Student gender: Female, n (%)",
  "Class type: Small class, n (%)",  "Area type, n (%)", "  Inner City", "  Rural",
  "  Suburban", "  Urban", "Class size, mean (SD)")
tab_df <- tab_df[, !names(tab_df) %in% "test"]
tab_df <- tab_df[, c("Variable", setdiff(names(tab_df), "Variable"))]
rownames(tab_df) <- NULL
knitr::kable(tab_df, caption = "Baseline characteristics by completeness status")
```

The clear baseline differences between complete and incomplete cases suggest that missingness is associated with observed covariates, indicating that the process is at least partly explainable by observed data. This makes the MAR assumption more plausible than MCAR. And relative to complete cases, incomplete observations are more likely to come from students receiving free lunch, from Black students, and from inner-city schools. In contrast, class type is relatively balanced between the two groups, and motivation shows almost no difference.

### Include strong auxiliary variables

```{r}
stars1 <- stars1 %>%
  arrange(student_id, grade) %>%
  group_by(student_id) %>%
  mutate(math_lag1 = lag(math_score, 1),
    math_lag2 = lag(math_score, 2),
    math_lag3 = lag(math_score, 3)) %>% ungroup()
cont_vars <- c("treads", "tlists", "wordsk", "motivr", "readbs", "mathbs", "math_lag1", "math_lag2", "math_lag3")

cor_df <- map_dfr(cont_vars, function(v) {
  tmp <- stars1 %>% select(math_score, all_of(v)) %>%
    filter(!is.na(math_score), !is.na(.data[[v]]))
  
  tibble(variable = v, n = nrow(tmp),
    correlation = cor(tmp$math_score, tmp[[v]], use = "complete.obs", method = "pearson"))
}) %>% arrange(desc(abs(correlation)))

cor_df
```

We examined the auxiliary variables including reading score (treads), word skills (wordsk), listening score (tlists), and lagged math scores. They have shown strong correlations with math achievement, which is informative for predicting missing math outcomes and can be included in the imputation model to reduce bias. This could be considered a case of satisfying the MAR assumption.

# Inferential Analysis

This section focuses on the final inferential model rather than the computational details used to prepare the data. Our objective is to estimate how initial assignment to a small kindergarten class affects the mathematics trajectory from kindergarten through Grade 3. Because the STAR data are longitudinal, clustered by school, and affected by nontrivial attrition, the logic of the inferential analysis must be made explicit. The strategy adopted here is therefore to state the required assumptions before each major modeling decision and to connect those decisions directly to the experimental design and the patterns seen in the EDA.

## Target estimand and treatment definition

Our primary estimand is the effect of initial kindergarten small-class assignment on K--3 mathematics outcomes. We deliberately use the kindergarten assignment indicator, rather than observed class type in every later grade, because the original randomization is cleanest at entry. By contrast, later observed class type is affected by mobility, reassignment, late entry, and the regular-versus-regular-aide restructuring documented in the STAR records and later analyses [@word1990star; @finn2007usersguide; @ding2006dynamic]. Under this design-based perspective, kindergarten assignment retains a defensible connection to the original experiment, while a model based on later observed class type would mix experimental assignment with post-randomization behavior.

This choice also clarifies the interpretation of the final model. We are not estimating the contemporaneous effect of class size in each later grade. Instead, we estimate how the initial kindergarten assignment is associated with the subsequent trajectory of math achievement, after adjustment for observed covariates and selective follow-up.

## Covariate completion before causal estimation

Before constructing the doubly robust estimator, we use multivariate imputation by chained equations (MICE) to complete incompletely observed covariates and auxiliary predictors that enter the retention and outcome models. This step is necessary because the doubly robust estimator requires a reasonably complete observed history for each student; otherwise, the propensity-to-remain model and the grade-specific outcome regressions would themselves be fitted on unstable or inconsistently restricted subsets of the data [@vanbuuren2011mice; @white2011multiple].

The role of MICE in this report is therefore not to replace the causal analysis, but to support it. We use fully conditional specification because the STAR data combine continuous and categorical variables, have irregular missingness across grades, and include predictors measured at multiple levels of the educational hierarchy. The imputation model includes treatment-related variables, baseline demographics, school identifiers, and auxiliary measures identified in the EDA as informative for later achievement. This is intended to make the conditional MAR assumption more plausible for the incomplete covariates used in the DR stage [@white2011multiple; @enders2010applied].

An important design choice is that MICE is used to recover covariate information needed for the attrition and outcome models, while the final inferential stage continues to treat outcome missingness as a substantive problem. In other words, the completed datasets are used to stabilize the observed covariate history, but the original missingness structure of the math outcomes is restored before the DR pseudo-outcome is constructed. This keeps selective follow-up explicit in the causal analysis rather than hiding it inside a fully imputed outcome panel.

The MICE step is used under the following working assumptions:

- **Conditional MAR for incomplete covariates**: after conditioning on the observed treatment indicators, baseline demographics, school information, and auxiliary predictors included in the imputation model, the missingness of incomplete covariates does not depend on their unobserved values.
- **Sufficiently rich conditional models**: the fully conditional specification is flexible enough to preserve the main relationships among treatment, covariates, and auxiliary variables that are needed later in the DR stage.
- **Respect for hierarchical structure**: because students are nested within schools, school-level information must be represented during imputation so that the completed covariate history is compatible with the multilevel outcome analysis.

Under these assumptions, the completed datasets can be used to stabilize the observed covariate history before causal estimation, while still leaving outcome attrition to be handled explicitly in the next stage.

```{r infer-stage1, include=FALSE}
imp_data <- readRDS("stars_mice_mids.rds")
stars1_wide <- readRDS("stars1_wide_final.rds")
n_imp <- imp_data$m
```

## Final longitudinal mixed model

### Limitations of complete-case analysis

The descriptive analysis suggests that complete-case inference would be difficult to justify. Missingness is structured across grades, baseline characteristics differ by completeness status, and several auxiliary variables are predictive of later achievement. These diagnostics do not prove MAR, and the EDA itself may contain imperfections, but they do make a purely MCAR explanation implausible. We therefore proceed under a working assumption that a substantial part of the missing-data process can be explained by observed student- and school-level information [@rubin1976inference; @white2011multiple; @finn2007usersguide]. Under that assumption, an analysis that ignores selective follow-up is less credible than one that adjusts for it directly.

### Doubly robust adjustment for selective follow-up

The pseudo-outcome $\widetilde{Y}_{ig}$ is introduced to reduce bias from selective follow-up. The EDA suggests that attrition and intermittent observation are related to observed disadvantage and prior achievement, so an ordinary regression on complete cases would be vulnerable to selection bias. Following the augmented inverse-probability-weighting framework of @robins1994estimation, we combine a model for retention with a model for the observed math outcome. The key appeal is the double-robust property: the treatment-effect estimator remains consistent if either the retention model or the outcome model is correctly specified, even if the other is not [@funk2011doubly; @seaman2013introduction]. In the STAR setting, where Grade 3 attrition can be substantial and is unlikely to be captured perfectly by a single parametric model, this is a more defensible strategy than relying only on OLS adjustment [@krueger1999experimental; @ding2006dynamic].

The DR step is used under the following working assumptions:

- **Conditional ignorability of outcome missingness**: after conditioning on the observed treatment history, baseline covariates, and auxiliary information carried into the completed datasets, whether a student remains observed is independent of the unobserved potential math outcomes.
- **Positivity / overlap**: every relevant subgroup has a non-negligible probability of remaining in the observed sample, so the inverse-probability weights are well defined and not dominated by extreme values.
- **Limited interference (SUTVA)**: one student's outcome can be interpreted with respect to their own initial assignment without being materially altered by the treatment status of other students.
- **Single-model protection through double robustness**: consistency of the DR estimator requires correct specification of at least one of the two nuisance models, namely the retention model or the outcome model [@robins1994estimation; @funk2011doubly].

These assumptions are standard in causal analyses with missing outcomes, but they remain substantive rather than automatic [@seaman2013introduction; @dashti2024tmle]. Under them, the DR pseudo-outcome reduces dependence on any one parametric model and provides a more credible basis for the final longitudinal analysis.

### Final model specification

Once the doubly robust pseudo-outcome is constructed, we summarize it with a longitudinal mixed-effects model. Let $\widetilde{Y}_{ig}$ denote the doubly robust pseudo-outcome for student $i$ at grade $g$, let $T_i$ denote the kindergarten small-class assignment indicator, let $X_i$ denote the baseline adjustment vector, let $s(i)$ index the student's baseline school, and let $b_i$ denote a student-specific random intercept. The final model is
$$
\widetilde{Y}_{ig} = \beta_0 + \beta_T T_i + \delta_g + \eta_g(T_i \times \text{Grade}_g) + \gamma^\top X_i + u_{s(i)} + b_i + \epsilon_{ig},
$$
where $\delta_g$ captures average grade-to-grade growth, $\eta_g$ allows the effect of kindergarten assignment to vary by grade, $u_{s(i)} \sim N(0, \sigma^2_{\text{school}})$ is a school-level random intercept, $b_i \sim N(0, \sigma^2_{\text{student}})$ is a student-level random intercept, and $\epsilon_{ig}$ is an idiosyncratic error term. This follows the repeated-measures logic of @laird1982random while adapting it to the clustered STAR design. With kindergarten as the reference grade, $\beta_T$ is interpreted as the small-class effect in kindergarten, and the grade-interaction terms describe how that initial effect changes in Grades 1, 2, and 3.

The final mixed model is interpreted under the following working assumptions:

- **Correct mean-structure specification**: the main systematic variation in the DR-adjusted outcome is adequately represented by initial treatment, grade, their interaction, and the selected adjustment covariates.
- **Appropriate dependence structure**: school-level and student-level random intercepts capture the dominant clustering induced by within-school assignment and repeated measurements on the same child.
- **Mean-zero residual variation**: conditional on the included fixed and random effects, the remaining residual component has mean zero and does not introduce systematic bias into the treatment coefficients.
- **Intent-to-treat style interpretation**: the fitted model is interpreted as a longitudinal analysis of initial assignment, not as a model of later contemporaneous class type after post-randomization switching.

Under these assumptions, the treatment main effect and treatment-by-grade interactions can be interpreted as summaries of how the initial kindergarten assignment relates to the subsequent K--3 achievement path. Two modeling choices deserve additional justification because they are central to the longitudinal interpretation: the multilevel random-effects structure and the treatment-by-grade interaction.

### Multilevel dependence and random intercepts

Project STAR randomized class type within schools rather than across the entire state, so students within the same school are expected to share common institutional environments and administrative features. The school random intercept $u_{s(i)}$ captures this shared school-level variation and respects the blocked nature of the original design [@word1990star; @finn2007usersguide]. At the same time, each child contributes repeated outcomes across grades, so observations from the same student are inherently correlated. The student random intercept $b_i$ accounts for persistent person-level heterogeneity in achievement and aligns the model with the repeated-measures framework of @laird1982random. In diagnostic checks across all 20 imputed datasets, this additional random effect did not produce singular fits or convergence failures, and its variance component remained stable, so retaining both levels of clustering is empirically defensible.

### Dynamic treatment effects across grades

The interaction between kindergarten assignment and grade is essential because the substantive question is not only whether small classes help on average, but whether the advantage persists, grows, or fades over time. Omitting the interaction would force the small-class effect to be constant from kindergarten through Grade 3. That is a strong restriction in a study like STAR, where prior work has explicitly examined cumulative and dynamic treatment effects over multiple years [@ding2006dynamic]. The interaction specification lets the data distinguish the initial kindergarten effect from subsequent changes in Grades 1, 2, and 3.

### Pooling and statistical inference

The same final mixed model is estimated in each of the `m = 20` completed datasets. Point estimates and confidence intervals are then pooled using Rubin's rules, so the reported uncertainty combines ordinary sampling variability with the extra uncertainty introduced by missing-data preprocessing [@rubin1987multiple]. All inferential results reported below therefore reflect both model-based uncertainty and uncertainty from missing-data handling, rather than treating any one completed dataset as if it were observed without error.

### Answering the research questions

The final model is intended to answer two connected substantive questions in a format parallel to the example report, but adapted to a pooled mixed-model setting. To keep the inferential logic transparent, the results are organized in the same order as the questions: first, a pooled joint test answers whether a small-class advantage exists at any point in K--3; second, a pooled interaction test evaluates whether that advantage changes across grades. The coefficient table and model-based plots are then used to interpret the magnitude and pattern of those answers.

Question 1 asks whether students assigned to small kindergarten classes achieve higher mathematics scores than students assigned to regular classes after adjustment for selection bias. Because the final model includes a treatment-by-grade interaction, this is not a single-coefficient question outside kindergarten. The relevant null hypothesis is therefore that the entire treatment block is zero,
$$
H_0:\ \beta_T = \eta_1 = \eta_2 = \eta_3 = 0,
$$
against the alternative that at least one of these treatment-related terms is nonzero. Substantively, rejecting this null means that initial assignment to a small class is associated with higher mathematics achievement at one or more points in the K--3 trajectory after adjustment for observed selection processes.

Question 2 asks whether the relationship between initial class assignment and mathematics achievement differs across grades K--3 after adjustment for selection bias. This is the question of effect heterogeneity over time. The null and alternative hypotheses are
$$
H_0:\ \eta_1 = \eta_2 = \eta_3 = 0
\qquad \text{versus} \qquad
H_A:\ \text{at least one } \eta_g \neq 0.
$$
Under the null, the small-class advantage is constant across grades, so the model reduces to a common treatment shift added to the grade profile. Under the alternative, the size of the treatment contrast changes as students progress from kindergarten to Grade 3.

For both questions, the decision rule is the conventional one used throughout this report: at significance level $\alpha = 0.05$, we reject the relevant null hypothesis when the pooled p-value from the joint test is below 0.05, and otherwise we fail to reject it. In a multiply imputed mixed model, these questions are most naturally evaluated through pooled joint tests of the treatment block and the interaction block rather than through a single-dataset likelihood-ratio test. Specifically, we use the `D1` pooled Wald test implemented in the `mice` framework, which combines fixed-effect estimates and their covariance matrices across the 20 imputed datasets to test nested sets of coefficients under Rubin-style multiple-imputation inference.

```{r infer-stage2, include=FALSE}
grades <- c('K', '1', '2', '3')
math_cols <- paste0("math_score_", grades)
dr_models_list <- vector("list", n_imp)
main_effect_models <- vector("list", n_imp)
null_effect_models <- vector("list", n_imp)
failed_idx <- integer(0)

for (i in seq_len(n_imp)) {
  current_prepared <- complete(imp_data, i)
  current_prepared$student_id <- stars1_wide$student_id
  current_prepared$stayed_g3 <- stars1_wide$stayed_g3

  for (s_id in c("school_id_K", "school_id_1", "school_id_2", "school_id_3")) {
    current_prepared[[s_id]] <- stars1_wide[[s_id]]
  }

  for (col in math_cols) {
    current_prepared[[col]] <- stars1_wide[[col]]
  }

  ps_fit <- glm(
    stayed_g3 ~ student_race + student_gender + birth_year + free_lunch_K + area_type_K,
    data = current_prepared,
    family = binomial()
  )
  pi_hat <- predict(ps_fit, type = "response")

  for (g in grades) {
    y_name <- paste0("math_score_", g)
    s_name <- paste0("school_id_", g)
    l_name <- paste0("free_lunch_", g)

    formula_out <- as.formula(
      paste(y_name, "~ class_type_K + student_race + student_gender +",
            l_name, "+ area_type_K + (1 |", s_name, ")")
    )

    out_fit <- lmer(formula_out, data = current_prepared[!is.na(current_prepared[[y_name]]), ])
    m_hat <- predict(out_fit, newdata = current_prepared, allow.new.levels = TRUE)

    y_obs <- current_prepared[[y_name]]
    r_ind <- as.numeric(!is.na(y_obs))

    current_prepared[[paste0("dr_math_", g)]] <-
      m_hat + (r_ind / pi_hat) * (tidyr::replace_na(y_obs, 0) - m_hat)
  }

  current_long <- current_prepared %>%
    pivot_longer(
      cols = starts_with("dr_math_"),
      names_to = "grade",
      names_prefix = "dr_math_",
      values_to = "dr_outcome"
    ) %>%
    mutate(
      grade = factor(grade, levels = grades),
      student_id = factor(student_id),
      school_id_K = factor(school_id_K)
    )

  fit_i <- tryCatch(
    lmer(
      dr_outcome ~ class_type_K * grade + student_race +
        free_lunch_K + (1 | school_id_K) + (1 | student_id),
      data = current_long
    ),
    error = function(e) e
  )

  if (inherits(fit_i, "error")) {
    failed_idx <- c(failed_idx, i)
    dr_models_list[[i]] <- NULL
  } else {
    dr_models_list[[i]] <- fit_i
    main_effect_models[[i]] <- lmer(
      dr_outcome ~ class_type_K + grade + student_race +
        free_lunch_K + (1 | school_id_K) + (1 | student_id),
      data = current_long
    )
    null_effect_models[[i]] <- lmer(
      dr_outcome ~ grade + student_race +
        free_lunch_K + (1 | school_id_K) + (1 | student_id),
      data = current_long
    )
  }
}
```

```{r infer-stage3, include=FALSE}
if (length(failed_idx) > 0) {
  stop("Stage 3 model failures: ", paste(failed_idx, collapse = ", "))
}

valid_models <- sapply(dr_models_list, function(x) inherits(x, "merMod"))
valid_main_models <- sapply(main_effect_models, function(x) inherits(x, "merMod"))
valid_null_models <- sapply(null_effect_models, function(x) inherits(x, "merMod"))
if (!all(valid_models) || !all(valid_main_models) || !all(valid_null_models)) {
  stop("Non-merMod entries found in one or more pooled model lists.")
}

mira_obj <- as.mira(dr_models_list)
final_pooled_dr <- pool(mira_obj)
pooled_summary <- summary(final_pooled_dr, conf.int = TRUE)

main_mira_obj <- as.mira(main_effect_models)
null_mira_obj <- as.mira(null_effect_models)

q2_test <- D1(mira_obj, main_mira_obj)
q1_test <- D1(mira_obj, null_mira_obj)

extract_d1_row <- function(test_obj) {
  res <- test_obj[["result"]]
  if (is.null(res) || nrow(res) < 1) {
    stop("D1 object does not contain a usable result matrix.")
  }
  c(
    stat = as.numeric(res[1, "F.value"]),
    df1 = as.numeric(res[1, "df1"]),
    df2 = as.numeric(res[1, "df2"]),
    p = as.numeric(res[1, "P(>F)"])
  )
}

q1_vals <- extract_d1_row(q1_test)
q2_vals <- extract_d1_row(q2_test)

if (!all(lengths(list(q1_vals, q2_vals)) == 4) || any(!is.finite(c(q1_vals, q2_vals)))) {
  stop("Unable to extract pooled joint-test results from D1 objects.")
}

hypothesis_tests <- tibble::tibble(
  Question = c(
    "Question 1: overall small-class advantage",
    "Question 2: grade-varying treatment effect"
  ),
  `Test Statistic` = c(q1_vals["stat"], q2_vals["stat"]),
  df1 = c(q1_vals["df1"], q2_vals["df1"]),
  df2 = c(q1_vals["df2"], q2_vals["df2"]),
  `p value` = c(q1_vals["p"], q2_vals["p"]),
  Decision = c(
    ifelse(q1_vals["p"] < 0.05, "Reject H0", "Fail to reject H0"),
    ifelse(q2_vals["p"] < 0.05, "Reject H0", "Fail to reject H0")
  )
)
```

```{r infer-stage3-hypothesis-tests}
hypothesis_tests %>%
  mutate(
    `Test Statistic` = round(`Test Statistic`, 3),
    df2 = round(df2, 2),
    `p value` = sapply(`p value`, fmt_p)
  ) %>%
  knitr::kable(caption = "Pooled joint tests for the two inferential questions")
```

The joint tests provide direct answers to the two research questions. Here the reported test statistic is the pooled `D1` Wald/F statistic for a nested block of coefficients, not a classical single-sample likelihood-ratio statistic. For Question 1, the pooled test of the full treatment block rejects the null hypothesis at the 5% level, indicating strong evidence that initial assignment to a small class is associated with higher mathematics achievement at some point in the K--3 period after adjustment for selection bias. For Question 2, the pooled interaction test does not cross the conventional 5% threshold, so we fail to reject the null hypothesis that the treatment effect is constant across grades. This does not imply that the interaction estimates are exactly zero; rather, it means that the observed grade-to-grade differences in the small-class advantage are not estimated with enough precision to conclude that the treatment effect truly varies over time.

### Coefficient-based interpretation

```{r infer-stage3-output}
pooled_summary %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = sapply(p.value, fmt_p),
    conf.low = round(conf.low, 3),
    conf.high = round(conf.high, 3)
  ) %>%
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value,
    `CI Lower` = conf.low,
    `CI Upper` = conf.high
  ) %>%
  knitr::kable(caption = "Pooled mixed-model estimates for the inferential analysis")
```

The primary quantity of interest in Table 1 is the pooled coefficient on `class_type_KSmall class`, which represents the estimated small-class advantage in kindergarten, the reference grade. The interaction terms `class_type_KSmall class:grade1`, `:grade2`, and `:grade3` indicate how that initial effect changes in later grades. In the current results, the kindergarten effect is clearly positive, while the interaction terms are comparatively small and imprecise. Taken together, this pattern supports a sustained positive advantage for students initially assigned to small classes, but it does not provide strong evidence that the effect systematically grows or shrinks over time.

The grade coefficients should also be interpreted carefully. Because kindergarten is the reference category, the coefficients for Grades 1, 2, and 3 represent average score gains on the original math-score scale relative to kindergarten. They are therefore expected to be much larger than the treatment coefficient, since they primarily reflect normal developmental progress across primary school rather than the incremental effect of class assignment. For example, the estimated Grade 3 coefficient is roughly 138.6 points above kindergarten, while the incremental small-class effect in kindergarten is about 7.6 points.

One useful way to contextualize the treatment effect is to compare it with the model-implied gain between adjacent grades on the same score scale. The estimated increase from Grade 2 to Grade 3 is about 38 points, so a kindergarten small-class advantage of roughly 7.6 points is about one-fifth of a typical one-year gain. This does not convert the treatment into "months of learning," but it does show that the estimated effect is nontrivial relative to annual academic growth in the sample.

The adjustment terms for race and free-lunch status are included to improve comparability across students, but they are not the substantive focus of this section. For that reason, the most important quantities to compare in Table 1 are the treatment main effect, the treatment-by-grade interactions, and the size of the grade coefficients relative to the treatment effect. Under the assumptions stated above, the final model is best interpreted as a longitudinal intent-to-treat style analysis of initial assignment rather than a model of contemporaneous class type in each later grade.

### Model-based visualization and grade-specific contrasts

```{r infer-stage3-plot, fig.width=7, fig.height=5}
pred_data <- expand.grid(
  class_type_K = c('Regular class', 'Small class'),
  grade = factor(c('K','1','2','3'), levels = c('K','1','2','3')),
  student_race = 'Non-Black',
  free_lunch_K = 'Non-free lunch'
)

betas <- pooled_summary$estimate
names(betas) <- pooled_summary$term

term_or_zero <- function(term) {
  if (term %in% names(betas)) betas[[term]] else 0
}

pred_data <- pred_data %>%
  mutate(
    grade_main = case_when(
      grade == 'K' ~ 0,
      grade == '1' ~ term_or_zero('grade1'),
      grade == '2' ~ term_or_zero('grade2'),
      grade == '3' ~ term_or_zero('grade3')
    ),
    treat_main = ifelse(class_type_K == 'Small class',
      term_or_zero('class_type_KSmall class'), 0),
    treat_grade = case_when(
      class_type_K == 'Small class' & grade == '1' ~ term_or_zero('class_type_KSmall class:grade1'),
      class_type_K == 'Small class' & grade == '2' ~ term_or_zero('class_type_KSmall class:grade2'),
      class_type_K == 'Small class' & grade == '3' ~ term_or_zero('class_type_KSmall class:grade3'),
      TRUE ~ 0
    ),
    covar_adj = term_or_zero('student_raceNon-Black') +
      term_or_zero('free_lunch_KNon-free lunch'),
    predicted = term_or_zero('(Intercept)') + grade_main +
      treat_main + treat_grade + covar_adj
  ) %>%
  select(class_type_K, grade, predicted)

effect_by_grade <- tibble(
  grade = factor(c('K', '1', '2', '3'), levels = c('K', '1', '2', '3')),
  small_class_effect = c(
    term_or_zero('class_type_KSmall class'),
    term_or_zero('class_type_KSmall class') + term_or_zero('class_type_KSmall class:grade1'),
    term_or_zero('class_type_KSmall class') + term_or_zero('class_type_KSmall class:grade2'),
    term_or_zero('class_type_KSmall class') + term_or_zero('class_type_KSmall class:grade3')
  )
)


ggplot(pred_data, aes(x = grade, y = predicted, group = class_type_K, color = class_type_K)) +
  geom_line(size = 1.1) +
  geom_point(size = 3) +
  labs(title = 'Predicted Math Trajectory by Kindergarten Assignment',
       subtitle = 'Reference covariate profile; treatment effect allowed to vary by grade',
       x = 'Grade', y = 'Predicted DR-adjusted Math Score', color = 'Kindergarten Assignment') +
  theme_minimal()
```

Figure 1 translates the regression coefficients into predicted achievement trajectories for a common reference profile, defined here as a non-Black student who is not eligible for free lunch. This standardization is useful because interaction models are often easier to understand visually than through coefficients alone. The vertical distance between the two lines at each grade represents the model-implied small-class advantage for that grade, while the slope within each line reflects the expected grade-to-grade academic growth on the DR-adjusted outcome scale.

The figure should be read as a descriptive summary of the fitted model rather than as raw sample means. Because the plot is constructed from pooled fixed-effect estimates, it isolates the treatment pattern after covariate adjustment and missing-data handling. A roughly parallel pattern would indicate a fairly stable treatment effect across grades, whereas visibly widening or narrowing separation would suggest that the advantage is changing over time. In the current results, the small-class trajectory stays above the regular-class trajectory throughout K--3, which is consistent with a persistent positive effect of initial small-class assignment.

```{r infer-stage3-effect-by-grade}
effect_by_grade %>%
  mutate(small_class_effect = round(small_class_effect, 3)) %>%
  rename(
    Grade = grade,
    `Estimated Small-Class Effect` = small_class_effect
  ) %>%
  knitr::kable(caption = "Implied small-class effect by grade from the interaction model")
```

Table 2 makes the interaction model easier to interpret by converting the treatment main effect and the treatment-by-grade interaction terms into a single estimated small-class contrast for each grade. This presentation is especially helpful because the coefficient table alone does not directly show the total treatment effect outside the reference grade. If the estimates in Table 2 remain positive across grades, the evidence supports a sustained achievement advantage for the initially assigned small-class group; if they were to move toward zero or change sign, that would indicate attenuation or reversal of the early treatment benefit. In the present analysis, the grade-specific contrasts remain positive in every grade, although the earlier coefficient table shows that the changes over time are estimated with limited precision.

# Model Diagnostics

## MICE Fills Quality and Convergence Diagnostics

The core requirement of multiple imputation is that the algorithm must converge. We verify this by checking the trace plots of 20 imputations.

```{r}
plot(imp_data, c("math_score_3", "free_lunch_3", "teacher_degree_3"))
```

As shown in the figure above, the 20 curves representing different imputation sequences are highly tangled with no obvious trend or separation, indicating that the MICE algorithm has reached a stationary distribution and successfully converged. This means that the 20 'parallel worlds' we generated are statistically reliable.

## Propensity Score Model Diagnosis: Positivity Check

Double robust estimation relies on the overlap of propensity scores. We must ensure that in each imputed dataset, the stayed group and the attrited group have sufficient overlap in the predicted scores.

```{r}
# generate ps_list
ps_list <- lapply(1:imp_data$m, function(i) {
  d <- mice::complete(imp_data, i)
  
  d$stayed_g3 <- stars1_wide$stayed_g3
  
  # Propensity Model
  fit <- glm(stayed_g3 ~ student_race + student_gender + free_lunch_K + area_type_K, 
             data = d, family = binomial)
  
  return(predict(fit, type = "response"))
})

# Demonstrate the overlap using the first interpolation set as an example
ps_df <- data.frame(
  ps = ps_list[[1]], 
  status = factor(stars1_wide$stayed_g3, levels = c(0, 1), labels = c("Attrited", "Stayed"))
)

# Overlap Plot
ggplot(ps_df, aes(x = ps, fill = status)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("#e74c3c", "#3498db")) +
  labs(title = "Propensity Score Overlap (Imputation 1)", 
       subtitle = "Positivity Assumption Check",
       x = "Estimated Probability of Staying (pi_hat)", 
       fill = "Status") +
  theme_minimal()
```
The overlap plot shows that the propensity scores of the two groups of students are highly overlapping in most areas, and there are no weights extremely close to 0 or 1. This validates the Positivity Assumption, meaning that students from each background group have a chance to remain in the study.

Our Double Robust adjustment does not overly rely on a few extreme samples and satisfies the basic premise of causal inference.

## Verify the normality assumption of random effects

Mixed models assume that these random effects follow a normal distribution with a mean of 0, and if there is a severe deviation, the estimation of standard errors may fail. So we need to verify the normality of $u_s(i)$ (school random intercept) and $b_i$ (student random intercept)

```{r}
# extract
random_effects <- ranef(dr_models_list[[1]])

# Student-level
qqnorm(random_effects$student_id[[1]], main = "Q-Q Plot: Student Random Effects")
qqline(random_effects$student_id[[1]], col = "red")

# School-level
qqnorm(random_effects$school_id_K[[1]], main = "Q-Q Plot: School Random Effects")
qqline(random_effects$school_id_K[[1]], col = "red")
```


## Final Mixed Effects Model Diagnosis: Residual Analysis

We perform residual diagnostics on the final longitudinal model to verify the basic assumptions of linear regression.

```{r resid-diag, echo=FALSE, fig.width=8, fig.height=4}
# Run diagnostics on the first model
par(mfrow = c(1, 2))
plot(dr_models_list[[1]], which = 1) # Residuals vs Fitted
qqnorm(resid(dr_models_list[[1]]))
qqline(resid(dr_models_list[[1]]))
```

The residual plot shows that the mean is basically stable around the zero line

The Q-Q plot of the residuals indicates a certain degree of heavy-tailedness, which is common in large-scale educational assessment data due to individual variations in student performance. However, according to the Central Limit Theorem, our inferential consistency remains intact given the large sample size ($n > 10,000$). The stability of the treatment effect across 20 multiple imputations further confirms that our conclusions are not driven by these distributional tails.
# Sensitivity Analysis

## Comparison to Other Methods

## Alternative sample definitions

## Alternative imputation specifications

## Pseudo-missing cross-validation for imputation quality

# Discussion

# Conclusion

# References
