---
title: "Final Project for STAR Project"
author: "Ruitian Liu, Zhiye Jiang, Tommy Yu, Yuyu Zhang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    fig_width: 8
    fig_height: 5
    fig_caption: true
fontsize: 11pt
bibliography: references.bib
geometry: margin=1in
---


```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Lora:ital,wght@0,400;0,600;1,400&family=Source+Code+Pro:wght@400;600&family=Inter:wght@300;400;500;600&display=swap');

body {
  font-family: 'Inter', sans-serif;
  font-size: 15px;
  line-height: 1.85;
  color: #2d2d2d;
  background-color: #fafaf8;
}


h1, h2, h3, h4 {
  font-family: 'Lora', Georgia, serif;
  font-weight: 600;
  letter-spacing: -0.02em;
  color: #1a1a1a;
}

h1.title {
  font-size: 2.4em;
  border-bottom: 3px solid #2563eb;
  padding-bottom: 0.3em;
  margin-bottom: 0.2em;
}

h1 { font-size: 1.9em; border-left: 4px solid #2563eb; padding-left: 0.6em; }
h2 { font-size: 1.45em; color: #374151; border-bottom: 1px solid #e5e7eb; padding-bottom: 0.2em; }
h3 { font-size: 1.15em; color: #4b5563; }

.author, .date {
  font-family: 'Inter', sans-serif;
  font-size: 0.95em;
  color: #6b7280;
  font-weight: 400;
}


p { margin-bottom: 1.1em; max-width: 760px; text-align: justify;}


pre, code {
  font-family: 'Source Code Pro', 'Courier New', monospace;
  font-size: 0.88em;
}

pre {
  background-color: #f1f5f9;
  border: 1px solid #e2e8f0;
  border-left: 4px solid #2563eb;
  border-radius: 6px;
  padding: 1em 1.2em;
}


code:not(pre code) {
  background: #eff6ff;
  color: #1d4ed8;
  padding: 0.15em 0.4em;
  border-radius: 3px;
  font-size: 0.85em;
}


table {
  border-collapse: collapse;
  width: 100%;
  margin: 1.5em 0;
  font-size: 0.92em;
}

thead tr { background-color: #1e3a5f; color: #ffffff; }
th { padding: 10px 14px; text-align: left; font-weight: 600; }
td { padding: 9px 14px; border-bottom: 1px solid #e5e7eb; }
tbody tr:nth-child(even) { background-color: #f8fafc; }
tbody tr:hover { background-color: #eff6ff; transition: background 0.15s; }


blockquote {
  border-left: 4px solid #93c5fd;
  background: #eff6ff;
  margin: 1.5em 0;
  padding: 0.8em 1.2em;
  border-radius: 0 6px 6px 0;
  color: #1e40af;
  font-style: italic;
}


#TOC {
  font-size: 0.88em;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  padding: 1em;
  background: #ffffff;
  box-shadow: 0 1px 4px rgba(0,0,0,0.06);
}

#TOC ul { padding-left: 1.2em; }
#TOC li { line-height: 1.7; }
#TOC a { color: #374151; text-decoration: none; }
#TOC a:hover { color: #2563eb; }

img { border-radius: 6px; box-shadow: 0 2px 12px rgba(0,0,0,0.08); }

hr { border: none; border-top: 2px solid #e5e7eb; margin: 2em 0; }

.main-container { max-width: 960px; padding-left: 60px; padding-right: 60px;}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
pretty_label <- function(x) {
  x <- as.character(x)
  dict <- c(
    area_type = "Area Type",
    class_type = "Class Type",
    teacher_degree = "Teacher's Degree",
    teacher_career = "Teacher's Career Stage",
    student_gender = "Student's Gender",
    teacher_gender = "Teacher's Gender",
    free_lunch = "Free Lunch Status",
    teacher_race = "Teacher's Race",
    student_race = "Student's Race",
    math_score = "Math Score",
    "class_type * grade" = "Class × Grade"
  )
  out <- ifelse(x %in% names(dict), dict[x], x)
  out <- gsub("_", " ", out)
  out <- trimws(out)
  out <- sapply(strsplit(out, " +"), function(w) {
    paste0(toupper(substring(w, 1, 1)), substring(w, 2), collapse = " ")
  })
  unname(out)
}

fmt_p <- function(p, digits = 3, threshold = 1e-3) {
  if (is.na(p)) return(NA_character_)
  if (p < threshold) return(paste0("< ", formatC(threshold, format = "f", digits = digits)))
  formatC(p, format = "f", digits = digits)
}

options(warn = -1)
library(tidyverse)
library(tableone)
library(mice)
library(lme4)
library(gt)
library(patchwork)
library(stringr)
library(dplyr)
library(tableone)
library(tidyr)
library(patchwork)
library(purrr)
library(rcompanion)
library(psych)
library(ggplot2)
library(dlookr)
library(modelsummary)
library(broom)
library(forcats)
library(ggplot2)
```

# Abstract

# Background and study design

Project STAR (Student/Teacher Achievement Ratio) was one of the most important large-scale randomized education experiments in the United States. It was launched by the Tennessee State Department of Education in the 1985–86 school year to study whether smaller classes in the early elementary grades improve student outcomes. The project began in 79 schools across 42 districts in Tennessee and eventually followed students over many years. The large sample and long follow-up period had made STAR an important study in education fields.

In the STAR experiment, students entering kindergarten were randomly assigned within schools to one of three classroom types: small classes (about 13–17 students), regular classes (about 22–25 students), and regular classes with a full-time teacher’s aide. Teachers were also randomly assigned to these class types. The original design intended for students to stay in their assigned treatment group after entering the study. 

Class size has long been a question for educators. **Smaller classes** may help young children by giving them more individual attention and improving classroom participation. The potential benefits of class size may be especially important in the **early grades**. However, reducing class size is expensive, since it requires more teachers, more classrooms and funding. This is a big challenge for underserved schools or districts, and had become an important equity concern in education. For this reason, a question arises: **whether the gains of small classes are worthy to justify the substantial costs of implementation**. [@schanzenbach2007star]

Students were intended to stay in the same class-type assignment throughout the whole process. So the study was began with randomization, and all assignments was conducted within schools, the design helped balance school-level differences across treatment groups. This can be a good background for interpreting casual effects.

Although STAR began with randomization, later studies have claimed that it is **not a completely randomized experiments**. Because the experiments are affected by factors such as student transfers, classroom reassignment, imperfect compliance, teacher changes, and missing follow-up data. When missing of the data is not at random, it can generate selection bias. Therefore, studies must address these post-randomization outcomes carefully.


## Some empirical complications violating randomization

- **Student mobility / transfers.**

@chetty2011classroom claimed that some students did not remain in their originally assigned classrooms or schools over time, and new entrants were also added in later grades. As a result, treatment exposure in later grades may differ from the initial assignment.

- **Noncompliance with original assignment.**

@krueger1999star shown that some students moved between class types after the initial random assignment. Therefore, observed class type in later grades would not always be consistent with original treatment assignment.

- **Teacher changes and classroom reshuffling.**

@chetty2011classroom shown that if students were exposed to different classroom environments over time, including new teachers in subsequent grades and changes in class composition. These post-randomization changes may affect achievement independently of the original class-size assignment.

- **Peer Effect.**

@cascio2007first argued that scores may depend also on the characteristics of classmates. Since classroom composition can affect learning outcomes, peer effects may operate alongside the direct effect of small classes and make interpretation of the treatment difficult.

## Important Violations of Randomization - Selection Bias

One persistent concern of STAR Project is that there was substantial sample loss in the later years of the study, and this attrition may not have been entirely random. More broadly, @hanushek1999star identified attrition, crossovers, and selective test taking as threats to the integrity of STAR’s original randomization, and @filges2018classsize note that these remain central concerns in later assessments of the experiment.

- **Attrition is related to academic performance or student disadvantage.**

@ding2006dynamic shown that attrition in Project STAR was not random with respect to student performance. **Students who later attrited had significantly lower kindergarten achievement**, and that among students on free lunch, those who left the sample scored significantly **lower in mathematics** than comparable students who remained. In terms of this, if lower-performing students are more likely to leave the sample, the observed students may no longer represent randomization.

- **Attrition is also related to treatment status.**

@krueger1999star shown that attrition from STAR was substantial. In a subsample with initial enrollment logs, 10.4% of students assigned to small classes were missing by the start of kindergarten, compared with 14.3% of those assigned to regular classes, suggesting that some students may have been withdrawn because they were not assigned to a small class. In addition, @ding2006dynamic show that the pattern of **attrition differed between class types within and across schools**. These findings supported that the observed small-class and regular-class groups in later grades are no longer simply the originally randomized groups, but the outcome of initial random assignment followed by differential attrition and movement across class types.

# Research Questions

## Assumptions

- **Stable Unit Treatment Value Assumption (SUTVA):**

A first assumption is no interference. Each classroom’s potential outcome is assumed to be independent of the treatment assigned to other classrooms. The second assumption is no hidden versions of treatment. Small class should not contain multiple hidden versions of treatment.

- **Positivity:**

Within the relevant strata of background characteristics such as grade, race, free lunch status, and school context, there should be sufficient overlap so that no subgroup is deterministically assigned to only one class type.

- **Conditional ignorability of attrition:**

Conditional on the observed baseline covariates, treatment status, prior achievement, and other auxiliary variables related to student performance and participation, whether a student remains observed is assumed to be independent of the missing potential math outcomes.

- **Random assignment at baseline:**

At study entry, students were randomly assigned within schools to class types, so baseline treatment assignment can be treated as independent of students’ pre-treatment characteristics and potential outcomes.

- **Exclusive Restriction:**

Conditional on the class type actually received, the original random assignment affects math achievement only through treatment receipt, not through any direct channel such as parental responses, administrative reassignment, compliance behavior, etc.

## Questions of Interests

- How robust are the estimated treatment effects under multiple missing-data imputation methods?

- Whether missingness in math outcomes can be explained by observed background and prior-achievement variables, making a MAR-type assumption more plausible.

- Whether students assigned to smaller classes achieve higher mathematics scores than those in regular classes and whether this relationship differs across grades K--3 after resolving selection bias.

# Exploratory Data Analysis

```{r load-data}
# Please ensure the file exists at this path
df <- read.table("STAR.tab", header = TRUE, sep = "\t")
```

  
```{r phase-1}
## Phase 1: Column Name Filtering and Cleanin
# Truncate column names to the first 8 characters; STAR dataset column names usually contain grade prefixes (e.g., gk, g1, etc.)
names(df) <- substr(names(df), 1, 8)
cols <- names(df)

# Use regular expressions to extract column names starting with gk, g1, g2, g3
m <- regexpr("^(gk|g1|g2|g3)(.*)$", cols, perl = TRUE)
prefix <- regmatches(cols, m)
suffix <- sub("^(gk|g1|g2|g3)", "", prefix)

# Count the frequency of each variable and filter variables that exist across all grades (appearing at least 4 times)
tab <- table(suffix)
valid_suffix <- names(tab[tab >= 4])

# Construct the list of column names to keep
keep_cols <- unlist(lapply(valid_suffix, function(s) paste0(c("gk","g1","g2","g3"), s)))

# Separate basic student information (assuming the first 6 columns are basic info) and teaching data corresponding to grades
student_info <- df[, 1:6, drop = FALSE]
df_keep <- df[, intersect(names(df), keep_cols), drop = FALSE]
star <- cbind(student_info, df_keep)
```

  
```{r phase-2}
## Phase 2: Data Reshaping (Wide to Long -> Long to Wide)
# Convert the dataset from wide format to long format, then back to wide format, to standardize the grade structure
stars <- star %>%
  pivot_longer(cols = matches("^(gk|g1|g2|g3)"),
               names_to = c("grade", "variable"),
               names_pattern = "^g([k123])(.*)$", # Capture grade (k, 1, 2, 3) and variable names
               values_to = "value") %>%
  pivot_wider(names_from = variable,
              values_from = value) %>%
  # Convert the grade column into an ordered factor
  mutate(grade = dplyr::recode(grade, k = "K"),
         grade = factor(grade, levels = c("K", "1", "2", "3"), ordered = FALSE))

```


## Data Cleaning and Preprocessing

The dataset downloaded from Harvard Dataverse was first examined. It contains 11,601 observations and 379 variables, with a total of 2,477,601 missing entries. Given its large size and substantial amount of missing data, it is necessary to retain only variables relevant to the analysis and to address missingness carefully.

```{r phase-3}
## Phase 3: Renaming and Variable Type Conversion
# Standardize variable names to make them more readable
stars <- stars %>% rename(
  student_id = stdntid, birth_month = birthmon, birth_day = birthday, 
  birth_year = birthyea, teaching_years = tyears, class_size = classs, 
  math_score = tmaths, school_id = schid, teacher_id = tchid)
stars_diag <- stars %>% rename(area_type = surban, class_type = classt, teacher_degree = thighd,
    teacher_career = tcaree, student_gender = gender, teacher_gender = tgen,
    free_lunch = freelu, teacher_race = trace, student_race = race)

stars_diag[c("area_type","class_type","teacher_degree","teacher_career",
             "student_gender","teacher_gender","free_lunch",
             "teacher_race","student_race")] <- 
  lapply(stars_diag[c("area_type","class_type","teacher_degree","teacher_career",
                      "student_gender","teacher_gender","free_lunch",
                      "teacher_race","student_race")], factor)

stars_diag %>% dplyr::filter(if_all(where(is.factor), ~ !is.na(.))) %>% diagnose_category()
```

  
Following the Project STAR codebook, we recoded several categorical variables and collapsed categories when the original coding was too sparse for stable analysis. In particular, student race and teacher race were simplified to Black vs. Non-Black, since the STAR documentation shows that racial categories other than White and Black were extremely rare in the data. Teacher degree was reduced to Below Master versus Master or above, because specialist and doctoral categories were infrequent relative to the dominant bachelor’s/master’s groups. Teacher career stage was also collapsed into Lower level and Upper level, since the original multi-level ladder structure was highly imbalanced in practice.


```{r phase-4}
## Phase 4: Feature Engineering and Variable Cleaning
# Use case_when to create more descriptive categorical variables and delete unnecessary raw variables
stars1 <- stars %>% mutate(
  # Area type encoding conversion
  area_type = factor(case_when(
    surban == 1 ~ "Inner City",
    surban == 2 ~ "Suburban",
    surban == 3 ~ "Rural",
    surban == 4 ~ "Urban",
    TRUE ~ NA_character_)),
  # Class type encoding conversion
  class_type = factor(classt, levels = c(1, 2, 3),
                      labels = c("Small class", "Regular class", "Regular + aide")),
  # Teacher degree level binning
  teacher_degree = factor(case_when(
    thighd %in% c(1, 2) ~ "Below Master",
    thighd %in% c(3, 4, 5, 6) ~ "Master or above",
    TRUE ~ NA_character_),
    levels = c("Below Master", "Master or above"), ordered = TRUE),
  # Teacher career stage binning
  teacher_career = factor(case_when(
    tcaree %in% c(1, 2, 3) ~ "Lower level",
    tcaree %in% c(4, 5, 6, 7) ~ "Upper level",
    TRUE ~ NA_character_),
    levels = c("Lower level", "Upper level"), ordered = TRUE),
  # Gender and free lunch encoding conversion
  student_gender = factor(gender, levels = c(1, 2),
                          labels = c("Male", "Female")),
  teacher_gender = factor(tgen, levels = c(1, 2),
                          labels = c("Male", "Female")),
  free_lunch = factor(freelu, levels = c(1, 2),
                      labels = c("Free lunch", "Non-free lunch")),
  # Racial binarization
  teacher_race = factor(case_when(
    trace == 2 ~ "Black",
    trace %in% c(1, 3, 4, 5, 6) ~ "Non-Black",
    TRUE ~ NA_character_),
    levels = c("Black", "Non-Black")),
  student_race = factor(case_when(
    race == 2 ~ "Black",
    race %in% c(1, 3, 4, 5, 6) ~ "Non-Black",
    TRUE ~ NA_character_),
    levels = c("Black", "Non-Black"))
) %>%
  # Delete the converted raw variables and analysis indicators that are no longer needed; here, we only keep the math scores for performance variables
  select(-surban, -classt, -thighd, -gender, -race, -tcaree, -tgen, -freelu, -trace, -starts_with("gk"))
```


## Identification of potential confounder

Let the variables be defined as follows:

- $Y$: Outcome variable — students’ **math scores**.
- $X$: Treatment variables — **class type, grade level, and their interaction (class type × grade)**.
- $Z$: Potential confounders — **school area type, free lunch status, student race, teacher race, and teacher degree, etc**.
- $A$: Ancillary variables — **treads (reading), tlists (listening), wordsk (word skills), motivr (motivation), selfco (self-concept), and lagged math scores from previous grades**.

Our objective is to estimate the effect of class size on students’ mathematics achievement in the STAR experiment. Although the initial assignment was randomized, factors such as classroom reassignment and student mobility may weaken the ideal experimental conditions. As a result, background characteristics may influence both classroom assignment and student outcomes.

A variable $Z$ acts as a confounder if it is associated with both the treatment $X$ (class type) and the outcome $Y$ (math score). Thus we examine three relationships:

- **$X \rightarrow Y$:** whether class type is associated with math achievement.  
- **$Z \rightarrow Y$:** whether background characteristics are associated with math achievement.
- **$Z \rightarrow X$ or $X \leftarrow Z \rightarrow Y$:** whether background characteristics are related to classroom assignment. 

If these relationships hold, failing to control for $Z$ may bias estimates of the effect of class size on student outcomes. Therefore, these variables are included as controls in the subsequent analysis.


### Verify **$X \rightarrow Y$:**

```{r}
# Plots and Graphs
library(ggplot2)
stars1 <- stars1 %>%
  mutate(class_type = case_when(
    class_type %in% c("Regular class", "Regular + aide") ~ "Regular",
    TRUE ~ class_type
  )) %>%
  mutate(class_type = as.factor(class_type))
ggplot(data=stars1, mapping=aes(y=math_score,x=grade,fill=class_type)) +
   geom_boxplot(na.rm=TRUE) + 
   xlab('Grade Level')+ ylab('Math Scores') + 
   ggtitle('Math Scores by Level and Class Type')
```

The boxplots show the distribution of students’ mathematics scores by grade level (K–3) and class type (small class/regular class). We can see that math scores increase steadily across grades. Also, students in small classes consistently exhibit slightly higher math scores than those in regular classes within the same grade. Thirdly, the difference between small classes and regular classes appears to persist across all grade levels, suggesting a potential interaction between grade and class type. In particular, the gap between the two class types appears to be slightly larger in lower grades. Thus, we can say that $X \rightarrow Y$ exists.


### Examine potential variables that **$Z \rightarrow Y$:**

```{r, width = 10, height = 10}
library(patchwork)
options(warn = -1)
cat_vars <- c("area_type", "class_type", "teacher_degree", "teacher_career",
              "student_gender", "teacher_gender", "free_lunch", "teacher_race", "student_race")
make_all_plots <- function(df, cont, cats, ncol = 3) {
  cont_lab <- pretty_label(cont)
  one_cat <- function(cat_var) {
    d <- dplyr::filter(df, !is.na(.data[[cat_var]]), !is.na(.data[[cont]]))
    if (nrow(d) == 0) return(NULL)

    p <- kruskal.test(stats::as.formula(paste(cont, "~", cat_var)), data = d)$p.value
    p_txt <- fmt_p(p)
    y <- stats::quantile(d[[cont]], c(.01, .99), na.rm = TRUE)
    cat_lab <- pretty_label(cat_var)

    p1 <- ggplot2::ggplot(d,
      ggplot2::aes(.data[[cat_var]], .data[[cont]], fill = .data[[cat_var]])) +
      ggplot2::geom_boxplot(outlier.alpha = .3, outlier.size = .8, na.rm = TRUE) +
      ggplot2::coord_cartesian(ylim = y) +
      ggplot2::labs(title = paste0(cat_lab, " (p = ", p_txt, ")"),
        x = NULL, y = cont_lab) +
      ggplot2::theme_bw(base_size = 9) +
      ggplot2::theme(legend.position = "none")

    p2 <- ggplot2::ggplot(d,
      ggplot2::aes(.data[[cont]], fill = .data[[cat_var]], color = .data[[cat_var]])) +
      ggplot2::geom_density(alpha = .35, linewidth = .6, na.rm = TRUE) +
      ggplot2::labs(x = cont_lab, y = "Density") +
      ggplot2::theme_bw(base_size = 9)+
      ggplot2::theme(legend.position = "none")
    patchwork::wrap_plots(p1, p2, ncol = 2, widths = c(1, 1.15))
  }

  plots <- Filter(Negate(is.null), lapply(cats, one_cat))

  patchwork::wrap_plots(plots, ncol = ncol) +
  patchwork::plot_annotation(
    title = paste0("Distributions of ", cont_lab, " under different settings"),
    subtitle = "Kruskal–Wallis test p-values are reported in panel titles",
    theme = ggplot2::theme(
      plot.title = ggplot2::element_text(hjust = 0.5),
      plot.subtitle = ggplot2::element_text(hjust = 0.5))
  )
}

make_all_plots(stars1, "math_score", cat_vars)
```

The figure compares math score distributions across several student, teacher, and school characteristics. As we can see, some of the distribution of data has a strong violations of normality. So we use Kruskal--Wallis test to evaluate whether math scores differ across groups. Unlike the ANOVA, Kruskal--Wallis it is a nonparametric method based on ranks, which means that there are no normality assumption of data.

The null hypothesis of Kruskal--Wallis is that the distributions of math scores are the same across all groups; equivalently, under a common-shape assumption, the group medians are equal. The alternative hypothesis is that at least one group differs from the others.

Because multiple tests are conducted across several covariates, we adopt a significance threshold of $\alpha =0.001$ to reduce the risk of false positives arising from multiple comparisons and the large sample size of the STAR dataset. The figure compares math score distributions across several student, teacher, and school characteristics. 

In this case, Kruksal--Wallis p--value of **area type, free lunch status, student race, teacher race, teacher gender, and teacher degree** are less or equal than 0.001. Thus, we reject the null hypothesis, showing that they are significant factors that influence students’ mathematics performance, indicating their relationship of $Z \rightarrow Y$. To determine whether they act as confounders, we further examine whether these variables are also associated with class type (or to say $Z \rightarrow X$ exists).


### Examine confounders that satistfies **$Z \leftarrow X \rightarrow Y$:**

```{r}

all_vars <- c("math_score", "class_type", "grade", "free_lunch", 
              "student_race", "area_type", "teacher_degree", "teacher_gender")

stars_clean <- stars1[complete.cases(stars1[, all_vars]), ]

model_1 <- lm(math_score ~ class_type * grade, data = stars_clean)
model_2 <- lm(math_score ~ class_type * grade + free_lunch, data = stars_clean)
model_3 <- lm(math_score ~ class_type * grade + free_lunch + area_type, data = stars_clean)
model_4 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type, data = stars_clean)
model_5 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type + teacher_degree, data = stars_clean)
model_6 <- lm(math_score ~ class_type * grade + free_lunch + student_race + area_type + teacher_degree + teacher_gender, data = stars_clean)

models <- list("M1" = model_1, "M2" = model_2, "M3" = model_3, "M4" = model_4, "M5" = model_5, "M6" = model_6)

model_note <- c("M1: Baseline", "M2: + Free lunch", "M3: + Student race",
  "M4: + Area type", "M5: + Teacher degree", "M6: + Teacher gender")

keep_terms <- c("class_typeSmall class", "class_typeSmall class:grade1",
  "class_typeSmall class:grade2", "class_typeSmall class:grade3",
  "grade1", "grade2", "grade3")

term_labels <- c(
  "class_typeSmall class" = "Small class",
  "class_typeSmall class:grade1" = "Grade 1 × Small class",
  "class_typeSmall class:grade2" = "Grade 2 × Small class",
  "class_typeSmall class:grade3" = "Grade 3 × Small class",
  "grade1" = "Grade 1",
  "grade2" = "Grade 2",
  "grade3" = "Grade 3"
)

coef_df <- imap_dfr(models, ~{
  tidy(.x, conf.int = TRUE) %>%
    mutate(model = .y)
}) %>%
  filter(term %in% keep_terms) %>%
  mutate(term_label = recode(term, !!!term_labels),
    term_group = case_when(term == "class_typeSmall class" ~ "Treatment main effect",
      str_detect(term, "class_typeSmall class:") ~ "Treatment × grade",
      str_detect(term, "^grade") ~ "Grade main effect"),
    model = factor(model, levels = c("M1", "M2", "M3", "M4", "M5", "M6"))) %>%
  group_by(term) %>%
  arrange(model, .by_group = TRUE) %>%
  mutate(pct_change_prev = 100 * (estimate - lag(estimate)) / abs(lag(estimate)),
    label_change = case_when(
      row_number() == 1 ~ "Base",
      is.infinite(pct_change_prev) | is.nan(pct_change_prev) ~ "",
      TRUE ~ sprintf("%+.1f%%", pct_change_prev))) %>%
  ungroup() %>%
  mutate(model = fct_rev(model))


coef_df <- coef_df %>%
  group_by(term_label) %>%
  mutate(xmin = min(conf.low), xmax = max(conf.high), xrange = xmax - xmin, label_x = estimate + 0.05 * xrange) %>% ungroup()


p <- ggplot(coef_df, aes(x = estimate, y = model, color = term_group)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
  geom_point(size = 2.4) +
  geom_text(aes(x = label_x, label = label_change), size = 3, hjust = 0, color = "black") +
  facet_wrap(~ term_label, scales = "free_x", ncol = 3) +
   labs(title = "Coefficient stability across nested models",
    subtitle = paste("Baseline model (M1): Math Score ~ Class Type * Grade",
      "Labels show % change relative to the previous model", sep = "\n"),
    x = "Coefficient estimate (95% CI)", y = NULL,
    caption = paste("M1: Class Type * Grade", "M2: + Free Lunch Status", "M3: + Student's Race",
      "M4: + Area Type", "M5: + Teacher's Degree", "M6: + Teacher's Gender", sep = "   |   "))+
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5)) +
  coord_cartesian(clip = "off")
p
```

Equivalent to the linear regression model

$$
Y_i=\alpha+X_i^\top\beta+Z_i^\top\gamma+\varepsilon_i,
$$

where $Y_i$ is math score, $X_i$ contains the treatment-related design terms for class type, grade, and their interaction terms, and $Z_i$ contains candidate confounders added sequentially across nested models. The figure tracks how the estimated coefficients $\beta$ change as additional covariates $Z_i$ are added step by step into the regression model. Each panel corresponds to one coefficient, each point is the estimate from a given model, and each horizontal line is its 95% confidence interval. The percentages show how much the coefficient changes relative to the previous model. 

The goal is to assess whether adding a covariate substantially changes the estimated effects of treatment variables. If a coefficient shifts noticeably after a variable is added, that variable may be acting as a confounder. As a heuristic, a covariate is flagged as a potential confounder if its inclusion changes a focal class-type coefficient by more than 10% in absolute value.

In this plot, the largest changes occur after adding free lunch status, area type and student’s race, while later adding teacher's degree and gender does not changes the coefficient of treatment effects. This suggests that these three variables account for the main confounding in the relationship of $Z \rightarrow X$ or adding $Z$ changes how $X \rightarrow Y$. Thus, in our final model we need to conditioning on the effects of free lunch status, area type and student’s race in order to accurately estimate the effect of treatment variable.


## Analysis of missing mechanism

- **MCAR (Missing Completely at Random)**: the probability of missingness is unrelated to both observed and unobserved data. In this case, missing values occur purely by chance, dropping incomplete cases does not introduce bias.

- **MAR (Missing at Random)**: the probability of missingness may depend on observed variables, but conditional on those observed variables, it does not depend on the missing value itself.

- **MNAR (Missing Not at Random)**: the probability of missingness still depends on the unobserved value itself, even after conditioning on observed data. This cannot be verified directly from the observed data alone, because the mechanism depends on values that are not observed.

In practice, these mechanisms cannot usually be distinguished with certainty from the observed data alone, especially in distinguishing MAR and MNAR. Therefore, rather than attempting to verify MAR directly, we assess its plausibility indirectly using the following diagnostic strategies:

- **Assess whether missingness is explainable by observed data**: model the missingness indicator using baseline covariates, lagged outcomes, and other observed variables. If missingness can be substantially explained by observed information, the MAR assumption becomes more plausible. [@rubin1976inference]

- **Include strong auxiliary variables**: add variables that are highly predictive of the missing outcome or of the missingness process into the imputation or likelihood-based model. This can reduce bias and strengthen the plausibility of MAR. [@collins2001comparison]; [@white2011multiple]

- **Check robustness under MAR-based methods**: compare results across approaches such as complete-case analysis, IPW, FIML, and MI under the same observed-data structure. Similar conclusions make MAR-based inference more credible. [@baraldi2010introduction]

- **Conduct MNAR sensitivity analysis**: evaluate how conclusions change under plausible departures from MAR using methods such as pattern-mixture models, selection models, or delta-adjusted imputation. Stable results suggest that the main findings are not overly sensitive to violations of MAR. [@daniels2000simple]


### Proving data is not MCAR

Each student’s outcome record is summarized by a four-digit indicator over grades K, 1, 2, 3, where 1 denotes an observed math score and 0 denotes a missing math score. For example, 1111 indicates complete observation across all four grades and 1001 indicates a non-monotone pattern with observation in K and Grade 3 but missingness in between.

We classified the K–3 observation patterns into five descriptive groups:

- Pattern 1111 was defined as **complete cases**. 
- Patterns 1110, 1100, 1000, 0100, and 0110 were grouped as **monotone attrition**, since once students entered a missing state, they did not subsequently return to full follow-up. 
- Patterns 0011 and 0001 were classified as **late entry**, as observations only began in later grades.
- Patterns 1010, 1001, 0101, 0010, and 1011 were classified as **intermittent**, reflecting non-monotone missingness with reappearance after one or more missing waves. 
- Pattern 0111 was separated as **no records in K**, because it there are a specific group of students who hasn't got kindergarten but observed continuously from Grade 1 onward.

```{r}
wide <- stars1 %>% mutate(obs = !is.na(math_score)) %>%
  select(student_id, grade, obs) %>%
  pivot_wider(names_from = grade, values_from = obs)

wide <- wide %>% mutate(K = ifelse(is.na(K), FALSE, K),
    `1` = ifelse(is.na(`1`), FALSE, `1`),
    `2` = ifelse(is.na(`2`), FALSE, `2`),
    `3` = ifelse(is.na(`3`), FALSE, `3`))

wide <- wide %>% mutate(pattern = paste0(K*1, `1`*1, `2`*1, `3`*1))

wide <- wide %>% filter(pattern != "0000") %>%
  mutate(complete = ifelse(pattern == "1111", "Complete", "Incomplete"))

stars_plot <- stars1 %>% left_join(wide %>% select(student_id, pattern), by = "student_id") %>%
  filter(!is.na(pattern), pattern != "0000", !is.na(math_score))

pattern_order <- stars_plot %>% distinct(student_id, pattern) %>% count(pattern, sort = TRUE) %>% pull(pattern)
pattern_order <- c("1111", setdiff(pattern_order, "1111"))

stars_plot2 <- stars1 %>%
  left_join(wide %>% select(student_id, pattern), by = "student_id") %>%
  filter(!is.na(pattern), pattern != "0000", !is.na(math_score)) %>%
  mutate(pattern_class = case_when(pattern == "1111" ~ "Complete cases",
      pattern %in% c("1110", "1100", "1000", "0100", "0110") ~ "Monotone attrition",
      pattern %in% c("0011", "0001") ~ "Late entry",
      pattern %in% c("1010", "1001", "0101", "0010", "1011") ~ "Intermittent",
      pattern %in% "0111" ~ "No records in K",
      TRUE ~ NA_character_)) %>%
  filter(!is.na(pattern_class))

class_count <- stars_plot2 %>% distinct(student_id, pattern_class) %>% count(pattern_class)

class_labels <- class_count %>% mutate(label = paste0(pattern_class, "\n(n=", n, ")")) %>%
  {setNames(.$label, .$pattern_class)}

stars_plot2 <- stars_plot2 %>%
  mutate(pattern_class = factor(pattern_class, 
                                levels = c("Complete cases","Monotone attrition", 
                                           "Late entry", "Intermittent", "No records in K")))

p3 <- ggplot(stars_plot2, aes(x = pattern_class, y = math_score, fill = pattern_class)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_x_discrete(labels = class_labels) +
  labs(title = "Math score distribution by missingness type",
    subtitle = "Scores pooled across grades K–3",
    x = "Missingness type",
    y = "Math score") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "none")
p3 
``` 

The boxplot shows clear differences in pooled math-score distributions across missingness types from kindergarten to Grade 3. The pooled math-score distributions suggest that complete cases, intermittent cases, have broadly similar achievement levels. However, **the monotone attrition group has lower math score while the late-entry group shows significant higher math score among all categories**. The pattern suggests that missingness type is related to student performance, which is inconsistent with a purely MCAR mechanism. Thus, missingness type itself is informative and should be analyzed specifically.

## Examine whether missingness can be explained by observed variables

```{r}
baseline_vars <- c("math_score", "free_lunch", "student_race",
  "student_gender", "class_type", "area_type","class_size")

stu_base <- stars1 %>% arrange(student_id, grade) %>%
  group_by(student_id) %>% slice(1) %>%
  ungroup() %>%
  select(student_id, all_of(baseline_vars)) %>%
  left_join(wide %>% select(student_id, complete), by = "student_id")

tab <- CreateTableOne(vars = baseline_vars, strata = "complete",
  data = stu_base, test = TRUE)
tab_mat <- print(tab, smd = TRUE, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
tab_df  <- as.data.frame(tab_mat)
tab_df$Variable <- c("n", "Math score, mean (SD)", "Free lunch: Non-free lunch, n (%)",
  "Student race: Non-Black, n (%)", "Student gender: Female, n (%)",
  "Class type: Small class, n (%)",  "Area type, n (%)", "  Inner City", "  Rural",
  "  Suburban", "  Urban", "Class size, mean (SD)")
tab_df <- tab_df[, !names(tab_df) %in% "test"]
tab_df <- tab_df[, c("Variable", setdiff(names(tab_df), "Variable"))]
rownames(tab_df) <- NULL
knitr::kable(tab_df, caption = "Baseline characteristics by completeness status")
```

The clear baseline differences between complete and incomplete cases suggest that missingness is associated with observed covariates, indicating that the process is at least partly explainable by observed data. This makes the MAR assumption more plausible than MCAR. And relative to complete cases, incomplete observations are more likely to come from students receiving free lunch, from Black students, and from inner-city schools. In contrast, class type is relatively balanced between the two groups, and motivation shows almost no difference.

### Include strong auxiliary variables

```{r}
stars1 <- stars1 %>%
  arrange(student_id, grade) %>%
  group_by(student_id) %>%
  mutate(math_lag1 = lag(math_score, 1),
    math_lag2 = lag(math_score, 2),
    math_lag3 = lag(math_score, 3)) %>% ungroup()
cont_vars <- c("treads", "tlists", "wordsk", "motivr", "readbs", "mathbs", "math_lag1", "math_lag2", "math_lag3")

cor_df <- map_dfr(cont_vars, function(v) {
  tmp <- stars1 %>% select(math_score, all_of(v)) %>%
    filter(!is.na(math_score), !is.na(.data[[v]]))
  
  tibble(variable = v, n = nrow(tmp),
    correlation = cor(tmp$math_score, tmp[[v]], use = "complete.obs", method = "pearson"))
}) %>% arrange(desc(abs(correlation)))

cor_df
```

We examined the auxiliary variables including reading score (treads), word skills (wordsk), listening score (tlists), and lagged math scores. They have shown strong correlations with math achievement, which is informative for predicting missing math outcomes and can be included in the imputation model to reduce bias. This could be considered a case of satisfying the MAR assumption.


# Inferential Analysis

# Sensitivity Analysis

## Comparison to Other Methods

## Alternative sample definitions

## Alternative imputation specifications

## Pseudo-missing cross-validation for imputation quality

## Discussion of MNAR risk

# Discussion

# Conclusion

# References