---
title: "Untitled"
author: "Zhiye Jiang"
date: "`r Sys.Date()`"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
# Data Preparation and Cleaning
### Load necessary libraries, covering data processing, string manipulation, visualization, and diagnostic analysis
library(stringr)
library(dplyr)
library(tableone)
library(tidyr)
library(patchwork)
library(purrr)
library(rcompanion)
library(psych)
library(ggplot2)
library(dlookr)
library(modelsummary)
library(broom)
library(forcats)
library(ggplot2)
library(WeightIt)
library(mice)
library(WeightIt)
library(MatchThem)
library(cobalt)
```

```{r}
# ====================================================================
# STEP 0: Load the Data
# ====================================================================
# Load the dataset exactly as you saved it from Phase 1
stars1 <- readRDS("stars_cleaned_eda.rds")

# ====================================================================
# STEP 1: PRE-PROCESS (LOCF on Long Data)
# ====================================================================
stars1_locf <- stars1 %>%
  # 1a. Sort chronologically for each student
  arrange(student_id, grade) %>%
  group_by(student_id) %>%
  # 1b. Safely fill fixed demographic traits backwards and forwards ACROSS ROWS
  fill(student_gender, student_race, birth_year, birth_month, .direction = "downup") %>%
  ungroup()

# ====================================================================
# STEP 2 REVISED: PIVOT TO WIDE FORMAT (Keeping ALL Confounders)
# ====================================================================
stars1_wide <- stars1_locf %>%
  # 1. Select absolutely everything, but drop the manual lags if you want 
  # (because Wide format natively creates lags, e.g., math_score_K IS the lag for Grade 1).
  # If you want to keep them, just leave them in the time-variant list.
  select(-math_lag1, -math_lag2, -math_lag3) %>% 
  
  pivot_wider(
    names_from = grade,
    # 2. Put ALL time-variant variables here so they split by grade
    values_from = c(
      math_score, class_type, class_size, area_type, free_lunch,
      school_id, teacher_id, teaching_years, teacher_degree, teacher_career, 
      teacher_gender, teacher_race, 
      treads, tlists, wordsk, motivr, selfco, readbs, mathbs
    ),
    names_glue = "{.value}_{grade}"
  ) %>%
  # 3. Create the missingness indicator
  mutate(stayed_g3 = as.numeric(!is.na(math_score_3))) %>%
  # 4. Factorize (You will need to add the new categorical variables to this list)
  mutate(across(c(student_gender, student_race, 
                  starts_with("class_type"), starts_with("free_lunch"), 
                  starts_with("area_type"), starts_with("teacher_gender"),
                  starts_with("teacher_race"), starts_with("teacher_degree"),
                  starts_with("teacher_career")), as.factor))

# Quick sanity check
cat("New Dimensions of Wide Data:\n")
dim(stars1_wide)
saveRDS(stars1_wide, "stars1_wide_final.rds")
```

```{r}
library(future)
# ====================================================================
# REVISED STEP 3: MICE Imputation with Cluster Awareness
# ====================================================================


# 1. Clean the input: Remove the constant variables flagged by mice
stars_mice_input <- stars1_wide %>%
  select(
    -student_id,       # Remove ID from the math
    -stayed_g3, 
    -starts_with("teacher_id"), 
    -starts_with("class_size"),
    -teacher_gender_K, # Flagged as constant
    -readbs_K,         # Flagged as constant
    -mathbs_K,          # Flagged as constant\
    # DROP THE ALIASES: We only need the baseline Area Type
    -area_type_1, -area_type_2, -area_type_3
  )

# Dry run with 0 iterations
ini <- mice(stars_mice_input, maxit = 0)

# See what method was assigned to each variable
# ini$method


# 2. Start with the sparse matrix (mincor = 0.15) to prevent singularity
pred_final <- quickpred(stars_mice_input, mincor = 0.15)

# 3. Identify our EDA "Must-Haves"
must_haves <- c("free_lunch_K", "free_lunch_1", "free_lunch_2", "free_lunch_3",
                "school_id_K", "school_id_1", "school_id_2", "school_id_3",
                "student_race", "area_type_K")

# 4. Identify our Target Outcomes (Math Scores)
math_outcomes <- c("math_score_K", "math_score_1", "math_score_2", "math_score_3")

# 4. FORCE the Must-Haves to predict the Math Outcomes
# This ensures that even if correlation was < 0.15 in a specific slice, 
# the structural and socio-economic context is always used.
pred_final[math_outcomes, must_haves] <- 1

# 5. Clean up: Ensure student_id is NEVER a predictor (Safety check)
if("student_id" %in% colnames(pred_final)) pred_final[, "student_id"] <- 0

# 3. Final Execution
imp_data <- futuremice(
  stars_mice_input,
  m = 20,
  maxit = 20,
  predictorMatrix = pred_final, # Using the thinned-out matrix
  ridge = 0.001,                  # Safety net for singularity
  parallelseed = 2026207
)

# 跑完后关闭并行池，释放后台资源（好习惯）
plan(sequential)

# Check if it succeeded!
cat("Parallel MICE successfully completed!\n")
```
Redundant Area/School Info: If area_type_K (Urban/Rural) is constant for every student within school_id_K, and you have both in the model, they are perfectly collinear.Constant Variables: Check if any column has the same value for almost everyone (e.g., if almost all students in your subset are the same race or have the same teacher degree).The "Wide Data" Trap: You have math_score_K, math_score_1, math_score_2, and math_score_3. If these are extremely highly correlated ($r > 0.95$), the matrix becomes unstable.

```{r}
# Extract one imputed dataset (e.g., the 1st one)
check_data <- complete(imp_data, 1)

# Check for missing values in your key columns
colSums(is.na(check_data[, c("free_lunch_3", "student_race", "school_id_K", "treads_3")]))

# Red lines = Imputed values; Blue line = Observed values
densityplot(imp_data, ~ math_score_3 + treads_3 + free_lunch_3)

```
```{r}
# Check the relationship between a covariate (e.g., Reading) and the outcome (Math)
xyplot(imp_data, math_score_3 ~ treads_3, pch = 18, cex = 0.7)


```

```{r}
# Compare the proportion of "Free Lunch" in observed vs imputed
prop.table(table(stars1_wide$free_lunch_3)) # Original
prop.table(table(check_data$free_lunch_3))   # Imputed

```

```{r}
# Plot convergence for your primary outcome and a key covariate
plot(imp_data, c("math_score_3", "free_lunch_3"))

pred_refined <- quickpred(stars_mice_input, mincor = 0.15)

pred_refined
```

```{r}
# ====================================================================
# STEP 4: 持久化保存 (Optimized Version)
# ====================================================================

cat("Saving results to disk...\n")

# 1. 保存 mids 对象 (用于诊断)
saveRDS(imp_data, "stars_mice_mids.rds")

# 2. 生成长表 (建议 include = FALSE，除非你需要对比原始缺失情况)
stars_imputed_long <- complete(imp_data, action = "long", include = FALSE) %>%
  # 关键：利用行号一一对应贴回 student_id
  # 只有当 stars_mice_input 的行序与 stars1_wide 完全一致时才有效
  mutate(student_id = rep(stars1_wide$student_id, times = imp_data$m)) %>%
  # 将 .imp 和 .id 移到前面方便观察
  select(student_id, .imp, .id, everything())

# 3. 保存长表 (用于后续分析)
saveRDS(stars_imputed_long, "stars_imputed_long.rds")

cat("Success! Total rows saved:", nrow(stars_imputed_long), "\n")
```

```{r}
# 加载画图需要的包
library(lattice)

# 检查一年级数学成绩的填补分布
densityplot(imp_data, ~ readbs_1)

# 检查一年级老师教龄的填补分布
densityplot(imp_data, ~ teaching_years_1)
```

```{r}
# 1. 提取填补后的数据集 (通常对 m 个数据集分别操作，这里以第 1 个为例)
completed_data <- complete(imp_data, 1)

# 2. 定义所有需要还原 NA 的成绩列
# 根据你的 Wide Format，我们需要还原 K 到 3 年级的所有成绩
math_cols <- c("math_score_K", "math_score_1", "math_score_2", "math_score_3")

# 3. 执行还原：只保留协变量的填补，成绩设回原始的 NA
stars_dr_prepared <- completed_data %>%
  mutate(
    student_id = stars1_wide$student_id,
    stayed_g3  = stars1_wide$stayed_g3
  ) %>%
  mutate(across(all_of(math_cols), ~ stars1_wide[[cur_column()]]))

# --- 快速检查 ---
cat("检查点 1: 数学成绩是否含有 NA?\n")
print(colSums(is.na(stars_dr_prepared[, math_cols]))) 
# 预期结果：这些列的 NA 数量应与 stars1_wide 完全一致

cat("\n检查点 2: 协变量（如 free_lunch_K）是否还有 NA?\n")
print(sum(is.na(stars_dr_prepared$free_lunch_K)))
# 预期结果：0 (已被 MICE 填满)
```
